# Rental spatial analysis

This section presents an analysis of the county’s existing rental supply, focusing on the amount and location of rental housing units. 

# Recent multifamily developments

Spatial analysis of recent multifamily developments; summarize general physical traits and location; talk about price points and contextualize with amenities or no amenities.


```{r recent}

#CoStar filter - Chesterfield County, VA; Property type: Multi-Family; Construction status: Existing; Rent Type: Market; Year Built Min: 2000

#Included Lat/Long in export

recent <- read_csv("data/costar_market_recent.csv") |> 
  clean_names()

```

# Market affordable rental

Spatial analysis of market affordable housing; summarize general physical traits and location; previously covered age and price point; pull in some information from previous but expand

```{r noah}

noah <- read_csv("data/chesterfield_noah_coded.csv") |> 
  clean_names()

```


# Affordable rental

Spatial analysis of affordable housing projects; describe projects, developer, subsidy, unit mix; status of subsidy; ask Jessica for list of state-funded projects in county


```{r affordable}

affordable <- read_csv("data/nhpd_chesterfield.csv") |> 
  clean_names() |> 
  filter(subsidy_status != "Inactive") |> 
  select(property_name, owner_name, owner_type, subsidy_name, subsidy_subname, assisted_units, known_total_units, lat, long, street_address)

```

# LIHTC

Low-income Housing Tax Credit projects are currently one of the major ways in which affordable housing is developed across the country. This program utilizes a tax incentive to 

```{r lihtc}

lihtc <- read_csv("data/nhpd_chesterfield.csv") |> 
  clean_names() |> 
  filter(subsidy_status != "Inactive") |> 
  select(property_name, owner_name, owner_type, subsidy_name, subsidy_subname, assisted_units, known_total_units, lat, long, street_address) |> 
  filter(subsidy_name == "LIHTC")

```