# (PART) COUNTYWIDE MARKET ANALYSIS {.unnumbered}

# Rental market summary {#part-2a-1}

This section summarizes the existing stock of rental units within the county and provides an analysis of recent price and production trends in the rental market.

```{r setup}

library(tidyverse)
library(tidycensus)
library(glue)
library(esquisse)
library(plotly)
library(scales)
library(formattable)
library(simplevis)
library(lubridate)
library(patchwork)
library(knitr)
library(zoo)
library(janitor)
library(qs)
library(readxl)
library(sf)
library(stringdist)
library(leaflet)
library(htmlwidgets)

```

## Rental housing stock

The proliferation of rental housing throughout Chesterfield County has been the result of affordability challenges across the region, as well as an increasing preference of [young professionals and baby boomers](https://www.freddiemac.com/research/consumer-research/20180404-preference-for-renting) for the flexibility offered by renting. This has resulted in strong demand for rental housing on a spectrum --- from luxury rentals to single room occupancy. 

### Total renters

Over the past decade, the number of renter households in the county has steadily been increasing. From 2010 to 2019, the number of renters in the county has gone from 24,312 to 30,273 --- a 24 percent increase. 

In 2020, there was a two percent decrease in renters. But this decrease of 633 renters may be attributed to [the collection challenges of the 2020 Census count](https://apnews.com/article/lifestyle-health-coronavirus-pandemic-census-2020-government-and-politics-9aebdfb9c1b834352676d4b95f2d9ea0) and [historical undercounting of renters](https://censuscounts.org/whats-at-stake/will-you-count-renters-in-the-2020-census/). As rising interest rates and home sales prices persist, the number of renters in the county can be expected to rise as well --- in addition to the need for a diverse mix of rental housing opportunities.

```{r tenure}

#Pull number of renter households across time from ACS Table B25003.

years <- 2010:2020

b25003_vars <- load_variables(2010, "acs5") |> 
  filter(str_sub(name, end = 7) %in% "B25003_")

b25003_raw <- map_dfr(years, function(yr){
  b25003_pull <- get_acs(
    geography = "county",
    county = "Chesterfield County",
    state = "VA",
    table = "B25003",
    year = yr,
    survey = "acs5",
    cache_table = TRUE
  ) %>%
    mutate(year = yr)
  })

b25003_vars_cleaned <- b25003_vars |> 
  separate(label, into = c("est", "total", "tenure"), sep = "!!") |> 
  select(variable = name, tenure) |> 
  drop_na() |> 
  mutate(tenure = case_when(
    tenure == "Owner occupied" ~ "Homeowner",
    tenure == "Renter occupied" ~ "Renter"
  )) |> 
  filter(tenure == "Renter")

b25003_data <- b25003_raw |> 
  right_join(b25003_vars_cleaned, by = "variable") |> 
  select(NAME, fips = GEOID, year, tenure, estimate, moe)
  
#Create line graph showing change in renter households across time.
renters_plot <- gg_line(
  b25003_data,
  x_var = year,
  y_var = estimate,
  y_breaks_n = 5,
  title = "Renter households",
  x_title = element_blank())

plotly::ggplotly(renters_plot) %>% 
  plotly_camera()

```

### Rental housing type

The rise in renters in Chesterfield County has not led to a commensurate rise in diverse housing options for renters. Over a third (37 percent) of rental housing in the county is made up of single-family detached homes --- that's an estimated 10,984 homes. The other third of rental housing is multifamily housing located in buildings with 5 to 19 units. These are typically low-rise and garden style apartments. The remaining third is a mix of smaller multifamily like duplexes, townhomes, and manufactured homes.

The distribution of rental housing type has largely remained the same since 2010. The major increases in rental housing have been among single-family detached home rentals (+ 2,502 since 2010) and multifamily buildings containing 5 to 9 units (+ 1,545). Although single-family rentals provide growing families with much needed space, multifamily development can more effectively address a rapidly growing demand at scale.

```{r structure}

years <- 2010:2020

b25032_vars <- load_variables(2010, "acs5") |> 
  filter(str_sub(name, end = 7) %in% "B25032_")

b25032_raw <- map_dfr(years, function(yr){
  b25032_pull <- get_acs(
    geography = "county",
    county = "Chesterfield County",
    state = "VA",
    table = "B25032",
    year = yr,
    survey = "acs5",
    cache_table = TRUE
  ) %>%
    mutate(year = yr)
  })

b25032_vars_cleaned <- b25032_vars |> 
  separate(label, into = c("est", "total", "tenure", "structure"), sep = "!!") |> 
  select(variable = name, tenure, structure) |> 
  drop_na() |> 
  filter(tenure == "Renter-occupied housing units")

b25032_data <- b25032_raw |> 
  right_join(b25032_vars_cleaned, by = "variable") |> 
  select(NAME, fips = GEOID, year, structure, estimate, moe)

structure_plot <- ggplot(
  b25032_data,
  aes(
    fill = fct_reorder(structure, estimate),
    y = estimate,
    x = year)) +
  geom_bar(position = "stack", stat = "identity")

plotly::ggplotly(structure_plot) %>% 
  plotly_camera()


```

#### Single-family rentals

Single-family home rentals (SFR) in the county have been on the rise --- experiencing a 28 percent increase in ten years, the most substantial percent growth among all rental housing types.

SFRs are a growing trend in the rental market as lifestyle changes and affordability challenges. The COVID-19 pandemic has also had a dramatic impact on the demand for SFRs, as many individuals and families looked avoid higher density apartments and had the ability to work and learn remotely.^[https://www.nmhc.org/research-insight/research-notes/2021/an-overview-of-single-family-rentals/]

Based on an analysis of owner address and parcel address, there are an estimated 13,205 single-family home rentals in the county. The majority of single-family rental owners in the county (61 percent) live somewhere else in the county, while 29 percent live somewhere else in Virginia. The remaining 10 percent of single-family rental owners live out-of-state.

```
THIS NEEDS FLESHED OUT A BIT MORE BASED ON THE SCRIPT BELOW. 

Maybe requires a better map that can be zoomed in on. I think there needs to be a bar chart showing distribution of owner location. And maybe it's worth doing a spatial join to Magisterial District to note where these are concentrated.

```


```{r sfr}

districts_geojson <- "https://services3.arcgis.com/TsynfzBSE6sXfoLq/ArcGIS/rest/services/Administrative/FeatureServer/9/query?outFields=*&where=1%3D1&f=geojson"

districts <- read_sf(districts_geojson)

plot(districts$geometry)

parcels <- st_read("data/ParcelsEnriched.shp")

parcels <- st_point_on_surface(parcels)

parcels_sub <- parcels %>%
  select(
    id = GPIN,
    use_code = UseCode,
    owner_address = OwnerAddre,
    owner_city = OwnerCity,
    owner_state = OwnerState,
    owner_zip = OwnerZip,
    address = Address,
    zip = Zip,
    year_built = Year_Built,
    improvement = Improvemen
  )

parcels_sub <- st_join(parcels_sub, left = FALSE, districts["MagDistName"])


parcels_sub$match_score <- stringsim(parcels_sub$owner_address, parcels_sub$address)

ggplot(parcels_sub, aes(x = match_score)) + 
  geom_histogram(bins = 30) + 
  theme_minimal()

set.seed(1983)

fuzzy_matches <- parcels_sub %>%
  filter(match_score >= 0.75 & match_score <= 0.9) %>%
  st_drop_geometry() %>%
  slice_sample(n = 100)

View(fuzzy_matches)

parcels_ownership <- parcels_sub %>%
  mutate(address_num = str_extract(address, "[0-9]+"),
         owner_num = str_extract(owner_address, "[0-9]+")) %>%
  mutate(ownership = case_when(
    use_code == "SD" & match_score > 0.6 & address_num == owner_num ~ "Owner-occupied SFH",
    use_code == "SD" ~ "Renter-occupied SFH",
    use_code %in% c("CD", "CN", "DU", "QP", "RE", "TH", "TN", "TP", "100") ~ "Multifamily",
    TRUE ~ "Other"
  ))

table(parcels_ownership$ownership)

library(tigris)
options(tigris_use_cache = TRUE)

ch_roads <- roads("VA", "Chesterfield", year = 2021)
ch_water <- area_water("VA", "Chesterfield", year = 2021)
ch_sfh <- filter(parcels_ownership, ownership %in% c("Owner-occupied SFH",
                                                     "Renter-occupied SFH"))

table(ch_sfh$improvement == 0)

sfh <- ggplot() + 
  geom_sf(data = ch_sfh, aes(fill = ownership), color = NA) + 
  geom_sf(data = ch_water, fill = "lightblue", color = NA) + 
  geom_sf(data = ch_roads, alpha = 0.5, color = "grey80") + 
  theme_void(base_size = 18) + 
  scale_fill_manual(values = c("#00CED1", "#e38520")) + 
  theme(legend.position = "bottom", plot.background = element_rect(fill = "#dcf5e0", color = NA),
        panel.background = element_rect(fill = "#dcf5e0", color = NA)) + 
  labs(fill = "")

plotly::ggplotly(sfh) %>% 
  plotly_camera()

sf_use_s2(FALSE)
zips <- zctas(cb = TRUE) %>%
  st_point_on_surface()
chesterfield <- counties("VA", cb = TRUE) %>%
  filter(NAME == "Chesterfield")

chesterfield_zips <- zips[chesterfield, ]

zip_list <- chesterfield_zips$GEOID20

rental_class <- ch_sfh %>%
  filter(ownership == "Renter-occupied SFH") %>%
  mutate(owner_location = case_when(
    owner_zip %in% zip_list ~ "Owner in Chesterfield",
    owner_state == "VA" ~  "Owner in Virginia",
    TRUE ~ "Owner out of state"
  ))

rental_pct <- rental_class %>%
  st_drop_geometry() %>%
  group_by(owner_location) %>%
  summarize(n = n()) %>%
  ungroup() %>%
  mutate(pct = 100 * (n / sum(n)))

sfr_bar <- gg_bar_col(rental_pct,
                      x_var = owner_location,
                      y_var = n,
                      col_var = owner_location)

plotly::ggplotly(sfr_bar) |> 
  plotly_camera()

```

The map below shows the spatial distribution of SFRs across the county. Much of the county's SFR is located between Chippenham Parkway (SR 150) and the World War II Veterans Memorial Highway (SR 288). This area is heavily single-family development from the last five decades. Additional densities of SFR can be seen along the Route One Corridor, as well as the Ettrick area of Chesterfield --- straddling Colonial Heights and Petersburg.

```{r sfr-map, fig.cap= "Map of SFR in the county"}

# sfr_pal <- colorFactor(c("green", "red", "blue"), domain = c("Owner in Chesterfield", "Owner in Virginia", "Owner out of state"))

# sfr_map <- leaflet(rental_class) |> 
#   addTiles() |> 
#   addCircleMarkers(
#     radius = 2,
#     color = ~sfr_pal(owner_location),
#     stroke = FALSE, fillOpacity = 0.5
#   )
# 
# saveWidget(sfr_map, file = "maps/sfr_map.html")

# ggplot() + 
#   geom_sf(data = rental_class, aes(fill = owner_location), color = "red") + 
#   geom_sf(data = ch_water, fill = "#8ecbe6", color = NA) + 
#   geom_sf(data = ch_roads, color = "grey50", lwd = 0.05) +
#   theme_void(base_size = 12) + 
#   scale_fill_manual(values = c("#15a19c", "#ba150f","#482d61"),
#                     labels = c("Chesterfield (61%)",
#                                "Virginia (29%)",
#                                "Out of state (10%)")) +
#   # theme(legend.position = "bottom", plot.background = element_rect(fill = "#dcf5e0", color = NA),
#   #       panel.background = element_rect(fill = "#dcf5e0", color = NA)) + 
#   labs(fill = "Rental Owner Location")


knitr::include_url("maps/sfr_map.html", height = "500px")

```


#### Manufactured homes

While single-family homes saw a major increase in rentals, manufactured homes experienced a significant decrease. From 791 in 2010 to 527 in 2020, the reported number of manufactured homes in the county has been cut down by a third.

This estimate from the U.S. Census Bureau's American Community Survey severely undercounts manufactured homes located in manufactured home communities.

Manufactured home communities are a valuable source of affordable housing across Virginia. Chesterfield County is home to a significant amount of MHCs --- about 1,500 in total. However, a significant issue faces manufactured home community residents. MHCs often give residents two options, 1) a resident may rent both their home and the lot on which that home is placed, or 2) a resident may own their home and rent the lot on which that home is placed.

In both cases, a resident is renting, but those residents in the second case are in an especially precarious situation should a MHC owner choose to sell or redevelop their property. MHCs are increasingly at-risk of redevelopment. In 2021, Suburban Village, along the Midlothian Turnpike, entered the news after a Maryland-based investment firm purchased the community for $23 million. As the county's third largest MHC with nearly 220 households, residents and advocates fear for substantial lot rent increases or displacement.

In contrast to growing investment firm interest in MHCs, local affordable housing nonprofit, project:HOMES, purchased a 52-unit MHC in 2021. Since the purchase, project:HOMES has made significant improvements to community infrastructure and has been working to replace distressed homes with new high-quality and energy efficient manufactured homes (with support from Virginia Tech's Virginia Center for Housing Research and Virginia Housing). project:HOMES's ultimate goal is to empower residents to take ownership of the community.


### Number of bedrooms

Most rental housing in the county are two-bedroom and three-bedroom homes. This accounts for nearly half of all rental housing in the county (68 percent). 

The number of four-bedroom rental homes has been on a steady increase since 2010 --- from 1,995 to 3,453 in 2020, a 73 percent increase. Although the increasing number of families in search of larger homes has led to a corresponding demand in larger rental homes, the impact can be felt among single adults, especially low-income adults, needing affordable rental opportunities. 

```{r bedrooms}

years <- 2010:2020

b25042_vars <- load_variables(2010, "acs5") |> 
  filter(str_sub(name, end = 6) %in% "B25042")

b25042_raw <- map_dfr(years, function(yr){
  b25042_pull <- get_acs(
    geography = "county",
    county = "Chesterfield County",
    state = "VA",
    table = "B25042",
    year = yr,
    survey = "acs5",
    cache_table = TRUE
  ) %>%
    mutate(year = yr)
  })

b25042_vars_cleaned <- b25042_vars |> 
  separate(label, into = c("est", "total", "tenure", "bedrooms"), sep = "!!") |> 
  select(variable = name, tenure, bedrooms) |> 
  drop_na() |> 
  filter(tenure == "Renter occupied")

b25042_data <- b25042_raw |> 
  right_join(b25042_vars_cleaned, by = "variable") |> 
  select(NAME, fips = GEOID, year, bedrooms, estimate, moe)

bedroom_plot <- gg_bar_col_facet(
  b25042_data,
  stack = TRUE,
  x_var = year,
  y_var = estimate,
  col_var = bedrooms,
  facet_var = bedrooms,
  y_breaks_n = 5,
  title = "Rental housing by bedrooms",
  x_title = element_blank()) +
    theme(
      legend.position = "none"
    )

plotly::ggplotly(bedroom_plot) %>% 
  plotly_camera()


```

### Rental housing age

The majority of existing rental housing (11,229 rental homes) in the county was built between 1960 and 1999. This represents a significant portion of rental housing that is at least 20 years of age. Aging rental homes often mean increasing maintenance costs or the need for significant renovations that get passed down to tenants. 



```{r yearbuilt}

b25036_vars <- load_variables(2020, "acs5") |> 
  filter(str_sub(name, end = 6) %in% "B25036")

b25036_raw <- get_acs(
    geography = "county",
    county = "Chesterfield County",
    state = "VA",
    table = "B25036",
    year = 2020,
    survey = "acs5",
    cache_table = TRUE
  )

b25036_vars_cleaned <- b25036_vars |> 
  separate(label, into = c("est", "total", "tenure", "yrbuilt"), sep = "!!") |> 
  select(variable = name, tenure, yrbuilt)

b25036_data <- b25036_raw |> 
  right_join(b25036_vars_cleaned, by = "variable") |> 
  select(NAME, fips = GEOID, yrbuilt, tenure, estimate, moe) |> 
  drop_na() |> 
  filter(tenure == "Renter occupied:") |> 
  mutate(yrbuilt = case_when(
    yrbuilt == "Built 2014 or later" ~ "2010 or later",
    yrbuilt == "Built 2010 to 2013" ~ "2010 or later",
    yrbuilt == "Built 2000 to 2009" ~ "2000 to 2009",
    yrbuilt == "Built 1990 to 1999" ~ "1980 to 1999",
    yrbuilt == "Built 1980 to 1989" ~ "1980 to 1999",
    yrbuilt == "Built 1970 to 1979" ~ "1960 to 1979",
    yrbuilt == "Built 1960 to 1969" ~ "1960 to 1979",
    yrbuilt == "Built 1950 to 1959" ~ "1940 to 1959",
    yrbuilt == "Built 1940 to 1949" ~ "1940 to 1959",
    yrbuilt == "Built 1939 or earlier" ~ "1939 or earlier"
  ))

built_plot <- gg_bar(
  b25036_data,
  x_var = yrbuilt,
  y_var = estimate,
  y_breaks_n = 5,
  title = "Rental housing by year built",
  x_title = element_blank()) +
    theme(
      legend.position = "none"
    )

plotly::ggplotly(built_plot) %>% 
  plotly_camera()

```

## Rental housing costs

The following data are from CoStar's multifamily database, which includes 118 existing rental properties in the county. Nine manufactured home communities are not found in this database; therefore, accurate and recent rent amounts for those properties are not available. Similarly, the single-family rental market is not tracked by CoStar, so these data do not include standalone homes for rent.

### Average asking rent

From the early 2000s, rental housing costs had generally been declining in the county. But changing trends in affordability and lifestyle have led to an increasing demand for rentals that has led to a complementary increase in rental housing costs. Since 2012, the average asking rent in the county has been on a steady increase. In Q3 2021, average asking rent in the county reached a two decade high of $1,510.

Average market asking rent per unit is expected to climb in the county --- signalling an increasing demand, as well as the construction of higher end rental apartments.

```{r costar-rents}

# This is pulling in average rents for all existing multifamily properties in the county. For some reason, this differs from the previous average rent calculations. I'd resolve this by updating the previous visualization in Part-1B-4.

#CoStar filter: Multi-family, Chesterfield County, VA; Construction Status: Existing

#Pulled quarterly Market Asking Rent Per Unit; Inflation Adjusted, No Forecast, History: All


rent <- read_csv("data/costar_rent_existing_may_12.csv") |>
  pivot_longer(cols = 2:3, names_to = "type", values_to = "value") |> 
  mutate(period = as.Date(as.yearqtr(period, format = "%Y Q%q"), frac = 1)) |> 
  mutate(type = str_replace_all(type, c("rent" = "Asking rent",
                                        "rent_adj" = "rent (inflation-adjusted)")))


costar_rent <-gg_line_col(
  rent,
  x_var = period,
  y_var = value,
  col_var = type,
  alpha_point = 0,
  size_line = 0.75,
  title = "Average asking rent in Chesterfield County",
  y_labels = label_dollar(),
  y_title = "Average asking rent",
  x_title = element_blank(),
  y_zero = TRUE,
  x_labels = scales::date_format(format = "%Y")
  ) +
  ggplot2::scale_x_date(limits = as.Date(c(NA, Sys.Date()))) +
  ggplot2::theme(
    legend.title = element_blank(),
    legend.position = "top",
    legend.direction = "horizontal",
    legend.justification = "center"
  )

plotly::ggplotly(costar_rent) %>% 
  plotly_camera()



```

### Rent by bedrooms

The number of bedrooms often impacts rental housing costs --- simply put, more bedrooms equals higher rent. But in Chesterfield County, studio apartments command a higher average asking rent than housing with more bedrooms. In Q1 2022, the average asking rent for a studio apartment was \$2,300, roughly \$600 more than a three-bedroom apartment. 

```

This section could use some work probably. I'm not sure if you want to address the studio rent prices or not. I'm not

```

```{r bedroom-rents}

#This uses the same filter as above from CoStar.


bedroom_rent <- read_csv("data/costar_br_existing_may_12.csv") |>
  pivot_longer(cols = 2:9, names_to = "type", values_to = "value") |>
  mutate(period = as.Date(as.yearqtr(period, format = "%Y Q%q"), frac = 1)) |> 
  mutate(rent = sub(".*_", "", type)) |> 
  mutate(rent = str_replace_all(rent, c("rent" = "Asking rent",
                                        "adj" = "rent (inflation-adjusted)")))
bedroom_rent <- bedroom_rent |> 
  filter(rent != "Asking rent") |> 
  mutate(type = substr(type,1,1))

costar_br_rent <-gg_line_col(
  bedroom_rent,
  x_var = period,
  y_var = value,
  col_var = type,
  alpha_point = 0,
  size_line = 0.75,
  title = "Average asking rent in Chesterfield County by bedroom",
  y_labels = label_dollar(),
  y_title = "Average asking rent",
  x_title = element_blank(),
  y_zero = TRUE,
  x_labels = scales::date_format(format = "%Y")
  ) +
  ggplot2::scale_x_date(limits = as.Date(c(NA, Sys.Date()))) +
  ggplot2::theme(
    legend.title = element_blank(),
    legend.position = "top",
    legend.direction = "horizontal",
    legend.justification = "center"
  )

plotly::ggplotly(costar_br_rent) %>% 
  plotly_camera()

```

## Rental housing construction

Like most of the country, the Great Recession negatively impacted home building in the county --- both in terms of single-family and multifamily housing. That impact was most felt among single-family home construction, however. Pre-recession, the average single-family residential construction permits numbered 2,050 and post-recession single-family home starts have been slow to recover to that pre-recession era. It was not until 2020 that the county exceeded that pre-recession average.

### Multifamily building permits

Multifamily (five or more units) home starts were not as impacted by the recession due to the generally low number of multifamily construction permits. Multifamily building permits saw a substantial increase in 2019 when it reached a two decade high of 1,466. 

Duplexes and smaller multifamily homes like tri-plexes and quads have not seen substantial construction in the county. These types of properties often make up what is called "[Missing Middle](https://missingmiddlehousing.com/)" housing, a term used to describe smaller multifamily and clustered single-family housing that often offers affordability by design (i.e., smaller homes typically mean lower price).

```
This could be fleshed out a lot more!!

```

```{r permits}

years <- 2000:2020

header_rows <- read_csv("https://www2.census.gov/econ/bps/County/co2020a.txt", 
                        col_names = FALSE,
                        n_max = 2)

column_names <- header_rows %>%
  select(X1:X18) %>%
  t() %>%
  as_tibble() %>%
  mutate(group = rep(1:6, each = 3)) %>%
  group_by(group) %>%
  fill(V1, .direction = "updown") %>%
  mutate(names = paste0(V1, ": ", V2)) %>%
  pull(names)

cbps_raw <- map_df(years, ~{
  raw <- read_csv(glue("https://www2.census.gov/econ/bps/County/co{.x}a.txt"), skip = 2, 
                    col_names = FALSE) %>%
    select(X1:X18) %>%
    set_names(column_names)
  
  raw
  
})

cbps_data <- cbps_raw %>% 
  mutate(year = `Survey: Date`,
         GEOID = paste0(`FIPS: State`, `FIPS: County`)) %>%
  select(`1-unit: Bldgs`:GEOID) %>%
  filter(GEOID == "51041") %>%
  pivot_longer(`1-unit: Bldgs`:`5+ units: Value`,
               names_to = "type",
               values_to = "value") %>%
  separate(type, into = c("Type", "col"), sep = ": ") %>%
  pivot_wider(names_from = col,
              values_from = value) %>%
  rename_with(tolower, Type:Value) %>% 
  select(GEOID, year, type:units)

permits <- gg_bar_col_facet(
  cbps_data,
  x_var = year,
  y_var = units,
  col_var = type,
  facet_var = type)

plotly::ggplotly(permits) |>  
  plotly_camera()

```

### Proposed and in-progress developments

There are currently 19 proposed or under construction multifamily developments in the county --- an estimated total of 5,039 rental homes^[Three properties do not have number of units listed and are slated for a 2027 completion date]. Six of these properties are expected to be completed in 2022, while the remaining homes will be completed over the next two years.

```
This could be fleshed out a lot more!!

The three properties slated for 2027 can probably be condensed into a single record. BizSense article refers to "150 low-rise multifamily units, 350 midrise units and 150 age-restricted units." Could maybe keep separate in this case. There are options!

```

```{r costar-construction}

under_con <- read_csv("data/costar_under_con_may_13.csv") |> 
  clean_names()

sum(under_con$number_of_units, na.rm = TRUE)


```

