# Rental spatial analysis

This section presents an analysis of the county’s existing rental supply, focusing on the amount and location of rental housing units. 

# Recent multifamily developments

Spatial analysis of recent multifamily developments; summarize general physical traits and location; talk about price points and contextualize with amenities or no amenities.

```{r setup}

library(sf)
```

```{r recent}

#CoStar filter - Chesterfield County, VA; Property type: Multi-Family; Construction status: Existing, Proposed, Under Construction; Rent Type: Market; Year Built Min: 2017

#Included Lat/Long in export

recent <- read_csv("data/costar_market_recent.csv") |> 
  clean_names()

#Bring in magisterial districts

districts_geojson <- "https://services3.arcgis.com/TsynfzBSE6sXfoLq/ArcGIS/rest/services/Administrative/FeatureServer/9/query?outFields=*&where=1%3D1&f=geojson"

districts <- read_sf(districts_geojson)

plot(districts$geometry)

recent <- as.data.frame(recent) |> 
  st_as_sf(coords = c("longitude", "latitude"), crs=4326, remove=FALSE)

newbuilds_by_district <- st_join(recent, left = FALSE, districts["MagDistName"])

leaf_sf_col(newbuilds_by_district,
            col_var = MagDistName)

```

# Market affordable rental

Spatial analysis of market affordable housing; summarize general physical traits and location; previously covered age and price point; pull in some information from previous but expand

```{r noah}

#Costar filter - Chesterfield County, VA, Property type: Multi-family; Construction status: Existing; Rent Type: Market; Class: B or C; CoStar Building Rating: three or fewer stars

#Include manufactured home communities identified by 2016 MHCCV study


noah <- read_csv("data/chesterfield_noah_coded.csv") |> 
  clean_names()

noah <- as.data.frame(noah) |> 
  st_as_sf(coords = c("longitude", "latitude"), crs=4326, remove=FALSE)

noah_by_district <- st_join(noah, left = FALSE, districts["MagDistName"])

leaf_sf_col(noah_by_district,
            col_var = MagDistName)


```


# Affordable rental

Spatial analysis of affordable housing projects; describe projects, developer, subsidy, unit mix; status of subsidy; ask Jessica for list of state-funded projects in county


```{r affordable}

affordable <- read_csv("data/nhpd_chesterfield.csv") |> 
  clean_names() |> 
  filter(subsidy_status != "Inactive") |> 
  select(property_name, owner_name, owner_type, subsidy_name, subsidy_subname, assisted_units, known_total_units, lat, long, street_address)

affordable <- as.data.frame(affordable) |> 
  st_as_sf(coords = c("long", "lat"), crs=4326, remove=FALSE)

affordable_by_district <- st_join(affordable, left = FALSE, districts["MagDistName"])


leaf_sf_col(affordable_by_district,
            col_var = MagDistName)

```

# LIHTC

Low-income Housing Tax Credit projects are currently one of the major ways in which affordable housing is developed across the country. This program utilizes a tax incentive to 

```{r lihtc}

lihtc <- read_csv("data/nhpd_chesterfield.csv") |> 
  clean_names() |> 
  filter(subsidy_status != "Inactive") |> 
  select(property_name, owner_name, owner_type, subsidy_name, subsidy_subname, assisted_units, known_total_units, lat, long, street_address) |> 
  filter(subsidy_name == "LIHTC")

lihtc <- as.data.frame(lihtc) |> 
  st_as_sf(coords = c("long", "lat"), crs=4326, remove=FALSE)

lihtc_by_district <- st_join(lihtc, left = FALSE, districts["MagDistName"])

leaf_sf_col(lihtc_by_district,
            col_var = MagDistName)

```