# Household income {#part-1b-2}

This section provides an analysis of household incomes in Chesterfield County disaggregated by tenure, age, and race/ethnicity. The major data source for this information will be the American Community Survey. Data will be provided at the county and census tract level.

## Background

In 2020, housing expenses increased amid the pandemic to account for nearly 35 percent of household spending — the largest share of average annual expenditures regardless of household type.

```{r global, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, 
                      message=FALSE, 
                      warning=FALSE, 
                      include=FALSE)

```

## Tenure

```{r tenure, message=FALSE, warning=FALSE}
library(tidyverse)
library(tidycensus)
library(readxl)

years <- 2010:2019

b25119_vars <- load_variables(2010, "acs5") %>%
  filter(str_sub(name, end = 6) %in% "B25119")

b25119_raw <- map_dfr(years, function(yr){
  b25119_pull <- get_acs(
    geography = "county",
    county = "Chesterfield County",
    state = "VA",
    table = "B25119",
    year = yr,
    survey = "acs1"
  ) %>%
    mutate(year = yr)
  })

b25119_vars_cleaned <- b25119_vars %>%
  separate(label, into = c("est", "income", "tenure"), sep = "!!") %>% 
  select(variable = name, tenure) %>%
  mutate(tenure = case_when(
    tenure == "Total" ~ "All households",
    tenure == "Owner occupied (dollars)" ~ "Homeowner",
    tenure == "Renter occupied (dollars)" ~ "Renter"
  ))

b25119_data <- b25119_raw %>%
  right_join(b25119_vars_cleaned, by = "variable") %>%
  select(NAME, year, tenure, estimate, moe) %>%
  group_by(year, tenure) %>%
  summarise(
    estimate = sum(estimate),
    moe = moe_sum(moe, estimate)
  ) %>% 
  ungroup() %>% 
  mutate(cv = ((moe/1.645)/estimate)*100) %>% 
  mutate(reliability = case_when(
    cv < 15 ~ "High reliability",
    cv >= 15 & cv <= 30 ~ "Medium reliability",
    cv > 30 ~ "Low reliability")
  )

cpi <- read_excel("data/CPI_U_RS.xlsx")
cpi <- cpi %>% 
  rename(year = Year,
         priceindex = Index) %>%
  transform(year = as.numeric(year))

b25119_cpi <- b25119_data %>%
    left_join(cpi, by = 'year') %>%
  transform(dollars19 = ((376.5/priceindex)*estimate)) %>%
  select(year, tenure, dollars19, moe, reliability)



# Reserved for Shiny app
```

## Age

```{r age}

years <- 2010:2019

b19049_vars <- load_variables(2010, "acs5") %>%
  filter(str_sub(name, end = 6) %in% "B19049")

b19049_raw <- map_dfr(years, function(yr){
  b25119_pull <- get_acs(
    geography = "county",
    county = "Chesterfield County",
    state = "VA",
    table = "B19049",
    year = yr,
    survey = "acs1"
  ) %>%
    mutate(year = yr)
  })

b19049_vars_cleaned <- b19049_vars %>%
  separate(label, into = c("est", "income", "age"), sep = "!!") %>% 
  select(variable = name, age) %>%
  mutate(age = case_when(
    age == "Total" ~ "All",
    age == "Householder under 25 years" ~ "Under 25 years",
    age == "Householder 25 to 44 years" ~ "25 to 44 years",
    age == "Householder 45 to 64 years"~ "45 to 64 years",
    age == "Householder 65 years and over"~ "65 years and over"
  ))

b19049_data <- b19049_raw %>%
  right_join(b19049_vars_cleaned, by = "variable") %>%
  select(NAME, year, age, estimate, moe) %>%
  group_by(year, age) %>%
  summarise(
    estimate = sum(estimate),
    moe = moe_sum(moe, estimate)
  ) %>% 
  ungroup() %>% 
  mutate(cv = ((moe/1.645)/estimate)*100) %>% 
  mutate(reliability = case_when(
    cv < 15 ~ "High reliability",
    cv >= 15 & cv <= 30 ~ "Medium reliability",
    cv > 30 ~ "Low reliability")
  )

cpi <- read_excel("data/CPI_U_RS.xlsx")
cpi <- cpi %>% 
  rename(year = Year,
         priceindex = Index) %>%
  transform(year = as.numeric(year))

b19049_cpi <- b19049_data %>%
    left_join(cpi, by = 'year') %>%
  transform(dollars19 = ((376.5/priceindex)*estimate)) %>%
  select(year, age, dollars19, moe, reliability)

# Reserved for Shiny app
```

## Race/ethnicity
```{r race}

years <- 2010:2019

b19013 <- c("B19013", paste0("B19013", LETTERS[2:9]))

concept_to_race <- function(x) {
  out <- x %>%
    str_remove_all("MEDIAN HOUSEHOLD INCOME IN THE PAST 12 MONTHS \\(\\IN 2019 INFLATION-ADJUSTED DOLLARS\\)\\ \\(|\\)") %>%
    str_to_title()
}

b19013_vars <- load_variables(2019, "acs5") %>%
  filter(str_sub(name, end = 6) %in% b19013) %>%
  filter(str_detect(name, "A") == FALSE) %>%
  mutate(race = concept_to_race(concept)) %>%
  mutate(race = case_when(
    race == "Median Household Income In The Past 12 Months (In 2019 Inflation-Adjusted Dollars" ~ "All",
    TRUE ~ race)) %>%
  mutate(across(.fns = ~str_remove_all(.x, "Householder")),
         across(.fn = ~str_remove_all(.x, "Alone"))) %>%
  mutate(race = case_when(
    race == "Two Or More Races " ~ "Multiracial",
    race == "White , Not Hispanic Or Latino " ~ "White, non-Hispanic",
    TRUE ~ race))
  

b19013_raw <- map_dfr(b19013, function(tb){
  yearly_data <- map_dfr(years, function(yr){
  b25119_pull <- get_acs(
    geography = "county",
    county = "Chesterfield County",
    state = "VA",
    table = tb,
    year = yr,
    survey = "acs1"
  ) %>%
    mutate(year = yr)
  })
  yearly_data
})

b19013_vars_cleaned <- b19013_vars %>%
  select(variable = name, race)

b19013_data <- b19013_raw %>%
  right_join(b19013_vars_cleaned, by = "variable") %>%
  select(NAME, year, race, estimate, moe) %>%
  group_by(year, race) %>%
  summarise(
    estimate = sum(estimate),
    moe = moe_sum(moe, estimate)
  ) %>% 
  ungroup() %>% 
  mutate(cv = ((moe/1.645)/estimate)*100) %>% 
  mutate(reliability = case_when(
    cv < 15 ~ "High reliability",
    cv >= 15 & cv <= 30 ~ "Medium reliability",
    cv > 30 ~ "Low reliability")
  )

cpi <- read_excel("data/CPI_U_RS.xlsx")
cpi <- cpi %>% 
  rename(year = Year,
         priceindex = Index) %>%
  transform(year = as.numeric(year))

b19013_cpi <- b19013_data %>%
    left_join(cpi, by = 'year') %>%
  transform(dollars19 = ((376.5/priceindex)*estimate))%>%
  select(year, race, dollars19, moe, reliability)

# Reserved for Shiny app
```

## Takeaways

[reserved]