# (PART) CORRIDOR CASE STUDIES {.unnumbered}

# Route 1 corridor analysis {#part-2b-1}

This section describes the Route 1 corridor in terms of its rental housing market, proximity of rental homes to amenities, recent major investments, and homeownership relative to the price of single-family rentals.

The Route 1 corridor used in this section uses the same six census tracts from the previous Route 1 affordability analysis in [Chapter 10](#part-1b-5).

```{r setup}

library(tidycensus)
library(tidyverse)
library(janitor)
library(leaflet)
library(sf)
library(sp)
library(tigris)
library(mapview)
library(simplevis)
library(readxl)
library(ggplot2)
library(lubridate)
library(viridis)
library(tidygeocoder)
library(knitr)
library(crsuggest)
options(tigris_use_cache = TRUE)
remotes::install_github("walkerke/mapboxapi")
library(mapboxapi)
library(rgdal)

```

## The Route One Corridor

U.S Route 1 once served as the major north-south corridor that connected East Coast states. However, the completion of Interstate 95 in the late 1950s drew traffic and business away from the highway. Since then, areas up and down the entirety of U.S. Route 1 have experienced decline and significant change that present challenges to revitalization. As a home to residents and businesses, the Route 1 corridor in Chesterfield County is well-positioned to be one of the county's most vibrant suburban communities to existing and future generations. In close proximity to the City of Richmond and with convenient access to Interstate 95, the corridor is centrally located in the Richmond metro area and could play a pivotal role in the growth and development of the entire region.

Current conditions on the northern section of Route 1 prevent significant reinvestment, but recent efforts by the County improve those conditions and incentivize new and much needed development. The 2018 Northern Jefferson Davis Special Area Plan and the 2019 Northern Jefferson Davis Real Estate Market Study provide Chesterfield County with initial tools to revitalize Route 1. Momentum has continued with the Route 1 Residential Overlay District, the Northern Jefferson Davis Incentive Program, and Greater Richmond Transit Company's Pilot Bus Service.

Much of the new rental housing development has occurred at the southern end of the study area --- pulled towards the intersection of W. Hundred Road (State Route 10) and Route 1, where there is a dense concentration of retail shopping options.

This study focuses its research on the six census tracts that straddle the Route 1 corridor from the Richmond-Chesterfield County border to the north and State Route 10 to the south. This encompasses portions of the Bensley, Bellwood, and northern Chester.

```{r route-one-map}

r1_tracts_list <- c("100300", "100404", "100405", "100406", "100407", "100409")

r1_tracts <- tracts("VA", "Chesterfield",
                    progress_bar = FALSE) |> 
  filter(TRACTCE %in% r1_tracts_list) |> 
  st_transform(4326) |> 
  mutate(neighborhood = case_when(
    TRACTCE == "100300" ~ "Bensley",
    TRACTCE == "100404" ~ "Bellwood",
    TRACTCE == "100405" ~ "Bensley",
    TRACTCE == "100406" ~ "Bellwood",
    TRACTCE == "100407" ~ "Bellwood",
    TRACTCE == "100409" ~ "Northern Chester"
  ))

hood_pal <- colorFactor(rainbow(3), c("Bensley", "Bellwood", "Northern Chester"))
  
leaflet(r1_tracts) |> 
  addTiles() |> 
  addPolygons(color = ~hood_pal(neighborhood), 
              weight = 3, 
              label = ~r1_tracts$NAMELSAD,
              labelOptions = labelOptions(noHide = F, 
                                          direction = "auto",
                                          textsize = "15px"))

```

## Rental housing supply

The following describes the existing rental housing supply along the Route 1 corridor.

### Number of rental homes

The number of rental homes along the Route 1 corridor has increased by 26 percent between 2010 and 2020 - a total of 708 renter households. Much of this rental housing growth occurred in the latter half of the decade --- from 2014 and on. From 2010 to 2012, rental housing was in decline --- most likely a reflection of the latter year impacts from the Great Recession. During this timeframe, the construction industry struggled to keep pace with the growing demand for new rental housing. From 2019 to 2020, there was a higher estimated percent change (13 percent)in renter homes than in previous years. Although this could be a significant impact of the COVID-19's impact on the rental housing market --- margins of error for the data in 2020 are also higher than previous years; and therefore, these changes should be interpreted with caution.

```{r r1-rhh}

years <- 2010:2020

b25003_vars <- load_variables(2020, "acs5") |> 
  filter(str_sub(name, end = 7) %in% "B25003_")

b25003_raw <- map_dfr(years, function(yr){
  b25003_pull <- get_acs(
    geography = "tract",
    county = "Chesterfield County",
    state = "VA",
    table = "B25003",
    year = yr,
    survey = "acs5",
    cache_table = TRUE,
    geometry = TRUE
  ) |> 
    mutate(year = yr)
  })

b25003_r1_raw <- b25003_raw |> 
  subset(GEOID %in% r1_tracts$GEOID)

b25003_vars_cleaned <- b25003_vars |> 
  separate(label, into = c("est", "total", "tenure"), sep = "!!") |> 
  select(variable = name, tenure) |> 
  drop_na() |> 
  mutate(tenure = case_when(
    tenure == "Owner occupied" ~ "Homeowner",
    tenure == "Renter occupied" ~ "Renter"
  )) |> 
  filter(tenure == "Renter")

b25003_r1 <- b25003_r1_raw |> 
  right_join(b25003_vars_cleaned, by = "variable") |> 
  select(NAME, fips = GEOID, year, tenure, estimate, moe)

b25003_route1 <- b25003_r1 |> 
  group_by(year) |> 
  summarise(estimate = sum(estimate))
  
#Create line graph showing change in renter households across time.
r1_renters_plot <- gg_line(
  b25003_route1,
  x_var = year,
  y_var = estimate,
  y_breaks_n = 5,
  title = "Renter households",
  x_title = element_blank())

plotly::ggplotly(r1_renters_plot) %>% 
  plotly_camera()

  
```

Based on 2010 and 2020 estimates, the increase in rental housing has largely occurred at the northern end of the Route 1 corridor in the Bensley area. All areas saw some increase in renter housing except for Census Tract 1004.04, which covers the eastern side of Route 1 from Falling Creek down to Bellwood Road. This census tract experienced a decrease of 180 renter households. While unable to determine the causes or from where these renter households came from, this area of Route 1 contains several manufactured home communities and naturally-occurring affordable housing --- all of which place households at greater risk of displacement.

```{r r1-rhh-map-10}

b25003_r1_10 <- b25003_r1 |> 
  filter(year == 2010) |> 
  select(NAME, fips, year, estimate)

b25003_r1_20 <- b25003_r1 |> 
  filter(year == 2020) |> 
  select(NAME, fips, year, estimate)

b25003_r1_10_20 <- cbind(b25003_r1_10, b25003_r1_20) |> 
  select(NAME, fips, est10 = estimate, est20 = estimate.1, geometry) |> 
  mutate(change = (est20-est10))



```

### Multifamily properties

Of the estimated 3,376 occupied rental homes from the U.S. Census Bureau, CoStar captures at most 2,417 of them that are located across 20 multifamily properties. These multifamily properties contain at least 12 rental homes and the majority of these properties (70 percent) are garden-style apartments. The majority (70 percent) of existing properties were built before 1976, and of these fourteen properties, only two have a recorded renovation --- Collindale Apartments (renovated in 1990) and Falling Creek (renovated in 2017).

```{r costar-mf-1, fig.cap= "Total number of multifamily units by property status"}

# Load .csv of CoStar properties. Data was collected by hand drawing boundaries for the necessary census tracts in the CoStar platform. No filters were placed on the search aside from multi-family property.

costar_r1_mf <- read_csv("data/costar_r1_mf.csv") |> 
  clean_names() |> 
  filter(property_address != "8219 Jefferson Davis Hwy") |> 
  filter(property_address != "9608 Jefferson Davis Hwy") |> 
  filter(property_address != "2537 Norcliff Rd") |> 
  filter(property_address != "12329-12331 Jefferson Davis Hwy") |> 
  filter(property_address != "9701 Jefferson Davis Hwy")

r1_unit_cnt <- costar_r1_mf |> 
  group_by(building_status) |> 
  summarise(units = sum(number_of_units))

r1_unit_cnt_bar <- gg_bar_col(
  r1_unit_cnt,
  x_var = building_status,
  y_var = units,
  col_var = building_status
)

plotly::ggplotly(r1_unit_cnt_bar) %>% 
  plotly_camera()

```

The city-county border was located much further north before the annexation of Northern Chesterfield by the City of Richmond in 1970. In spite of this change, these older rental properties represent the extent to which rental housing development in the mid-20th century stretched into the county.

```{r costar-mf-2, fig.cap="Map of multifamily properties along Route One Corridor"}

r1_map <- costar_r1_mf |> 
  st_as_sf(coords = c("longitude", "latitude"), crs=4326, remove=FALSE)

leaflet(r1_map) |> 
  addTiles() |> 
  addCircleMarkers(
    radius = 3) |> 
  addPolygons(data=r1_tracts, weight = 2, fillColor = NULL)


```

### Manufactured Home Communities

Manufactured housing grew out of the "travel trailer" popularity of the mid-20th century. These trailers evolved into the manufactured housing when the U.S. Department of Housing and Urban Development developed standardized building codes and the the Federal Housing Administration began offering mortgage insurance on both homes and lots.

Communities of manufactured homes grew out of the decline of "travel trailer" camp grounds and the increase in manufactured housing production. With the need for land to place manufactured homes, there were few options aside from existing land where similar types of housing were already permitted, or where a household could afford a plot of land themselves. Land owners also saw a value in developing manufactured home communities for those who wanted the experience of a single-family detached home, but could not afford to purchase.

But the stigma of manufactured housing has plagued both residents and developers for decades. There have been no new manufactured home communities in Virginia in that same amount of time. Nonetheless, manufactured home communities have carved out a particular niche along the housing spectrum, offering an affordable albeit often precarious housing opportunity.

Manufactured home communities offer rental housing opportunities across the county by one of two ways:

1.  A household owns their manufactured home and rents the land in which their home is placed (i.e. paying lot rent); or
2.  A household rents the manufactured home that is owned by the community owner (i.e. paying rent where lot rent is included).

Accurate data on manufactured home communities is difficult to come by. Community owners may allow for the additional of homes on a continuous basis without approval, or some homes may be left abandoned after a tenant has vacated. Anecdotally, some owners have even subdivided a single home into four units to be rented. This makes it difficult to have an accurate count of households.

The northern Route 1 corridor has a substantial number of manufactured home communities --- specifically 10 out of the county's known 17 manufactured home communities are located in the area. This accounts for at least 514 individual manufactured homes, or 31 percent of all homes located in a manufactured home community. This includes the sixth largest manufactured home park in the county, Shady Hill, which was recently purchased in July 2022.

```{r mhc-r1}

# Use MHCCV study instead; supplement additional MHCs.

mhc <- read_csv("data/mhccv_chesterfield.csv") 

mhc <- mhc |> 
  transform(fulladdress = paste(mhc$address, mhc$county, mhc$state, mhc$zip, sep = ", "))|> 
  select(name, fulladdress, units, units_mhccv, corridor)

mhc_coded <- mhc |> 
  geocode(address = fulladdress,
          method = 'geocodio',
          full_results = TRUE,
          unique_only = FALSE,
          lat = Latitude,
          long = Longitude)

mhc_coded <- mhc_coded |> 
  mutate(Latitude = case_when(
    name == "Holiday Mobile Home Park" ~ 37.393355,
    TRUE ~ Latitude
  )) |> 
  mutate(Longitude = case_when(
    name == "Holiday Mobile Home Park" ~ -77.427561,
    TRUE ~ Longitude
  ))

mhc_coded_r1 <- mhc_coded |> 
  filter(corridor == "Route 1 - N")


leaflet(mhc_coded_r1) |> 
  addTiles() |> 
  addCircleMarkers(
    radius = ~ (units_mhccv/5),
    color = "orange",
    stroke = FALSE, fillOpacity = 0.5
  )

sum(mhc_coded_r1$units_mhccv)

sum(mhc$units_mhccv)

```

Notable among these communities is Bermuda Estates, which was purchased by affordable housing nonprofit, project:HOMES, in September 2020. Only one of two nonprofit owned and managed manufactured home communities in the Commonwealth, Bermuda Estates has seen signficant infrastructure improvements and the replacement of homes with new energy efficient manufactured home models by project:HOMES.

```{r r1-mhc-list}

mhc_coded_r1 |> 
  arrange(desc(units_mhccv)) |> 
  select(1,4) |> 
  kable(caption = "Manufactured home communities along northern Route 1 corridor",
        align = "lr",
        col.names = c("Name", "Estimated homes")) %>% 
  kableExtra::kable_styling(
    bootstrap_options = c("condensed", "hover", "striped")) %>% 
  kableExtra::row_spec(8, bold=TRUE)

```

### Single-Family Rentals (SFR)

As mentioned previously in [Chapter 11](#part-2a-1), the growth of the single-family home rental (SFR) market is impacting all parts of the county. And the Route 1 corridor is no exception.

Along the northern Route 1 corridor, there is an estimated 1,008 single-family home rentals --- about 20 percent of all single-family homes the area.[^part-2b-1-1] The highest concentration of SFR is within Census Tract 1003 (Bensley), where 28 percent of all single-family homes are believed to be renter occupied.

[^part-2b-1-1]: This analysis was conducted by matching owner mailing address to physical address for all parcels wherein the land use code is "Single Dwelling." This analysis does not account for condominiums and therefore could be considered an undercount.

SFR is a growing demand among renters as homeownership continues to be out of reach for many, but it also serves a growing demand among households looking for greater space and privacy not offered in apartments. As seen in [Chapter 13](#part-2a-3), nearly 54 percent of recent multifamily housing is two-bedrooms or less, leaving many larger households competing for less supply. SFR is filling the gap where multifamily developers are often falling short due to increasing construction costs and public opposition.

```{r sfr-r1}

# Choropleth map of percent of SFR by census tract. Perhaps include a chart as well. Might be interesting to compare to change in homeownership rate for each.

sfr_r1 <- ch_sfh |> 
  st_join(left = FALSE, r1_tracts["NAMELSAD"]) |> 
  select(address, year_built, improvement, ownership, owner_zip, owner_state, NAMELSAD, geometry) |> 
  filter(improvement > 0)

sfr_r1_summary <- sfr_r1 |> 
  group_by(NAMELSAD, ownership) |> 
  summarise(count = n_distinct(address)) |> 
  mutate(total_sfh = case_when(
    NAMELSAD == NAMELSAD ~ sum(count))
  ) |> 
  mutate(percent_sfh = count/total_sfh) |> 
  filter(ownership == "Renter-occupied SFH") |> 
  st_drop_geometry()

sfr_r1_tracts <- r1_tracts |> 
  left_join(sfr_r1_summary, by = "NAMELSAD")

leaf_sf_col(sfr_r1_tracts,
            col_var = percent_sfh)

sum(sfr_r1_summary$count)

sum(sfr_r1_tracts$total_sfh)
```

The majority (68 percent) of owners of SFR along the corridor live somewhere else in the county, while 25 percent live outside of the county but somewhere else in Virginia. Only 7 percent of SFR owners live outside of Virginia --- the largest share of which has an owner in Florida.

```{r sfr_r1_owners}

sf_use_s2(FALSE)
zips <- zctas(cb = TRUE) %>%
  st_point_on_surface()
chesterfield <- counties("VA", cb = TRUE) %>%
  filter(NAME == "Chesterfield")

chesterfield_zips <- zips[chesterfield, ]

zip_list <- chesterfield_zips$GEOID20

# There is a duplicate address that I end up filtering out with a distinct count.

rental_class_r1 <- sfr_r1|> 
  filter(ownership == "Renter-occupied SFH") %>%
  mutate(owner_location = case_when(
    owner_zip %in% zip_list ~ "Owner in Chesterfield",
    owner_state == "VA" ~  "Owner in Virginia",
    TRUE ~ "Owner out of state"
  ))

rental_class_r1_summary <- rental_class_r1 |> 
  group_by(owner_location) |> 
  summarise(count = n_distinct(address))


```

### Rental prices

Median rents in the northern Route 1 corridor have fluctuated in the last decade. But increases in the typical rent have most significantly occurred in the southern part of Bellwood (Census Tract 1004.07) where Colony Village was developed. In this part of the corridor, median rent has increased by 24 percent from 2010 to 2020.

```{r acs-r1-rent}

years <- 2010:2020

b25064_vars <- load_variables(2020, "acs5") |> 
  filter(str_sub(name, end =6) %in% "B25064")

b25064_raw <- map_dfr(years, function(yr){
  b25064_pull <- get_acs(
    geography = "tract",
    county = "Chesterfield County",
    state = "VA",
    table = "B25064",
    year = yr,
    survey = "acs5",
    cache_table = TRUE,
    geometry = TRUE
  ) %>%
    mutate(year = yr)
  })

b25064_r1_raw <- b25064_raw |> 
  subset(GEOID %in% r1_tracts$GEOID)

cpi <- read_excel("data/CPI_U_RS.xlsx")
cpi <- cpi |>  
  rename(year = Year,
         priceindex = Index) |> 
  transform(year = as.numeric(year))

r1_rent_cpi <- b25064_r1_raw |> 
    left_join(cpi, by = 'year') |> 
  transform(dollars20 = ((381.2/priceindex)*estimate)) |> 
  select(GEOID, NAME, year, dollars20) |> 
  separate(NAME, into = c("NAMELSAD", "county", "state"), sep = ",") |> 
  transform(year_char = as.character(year))

r1_rent_change <- r1_rent_cpi |> 
  select(NAMELSAD, year, dollars20) |> 
  st_drop_geometry() |> 
  pivot_wider(
    names_from = year,
    values_from = dollars20
  )

r1_rent_change_map <- r1_tracts |>
  left_join(r1_rent_change, by = "NAMELSAD") |> 
  clean_names() |> 
  select(tract = namelsad, x2010:x2020) |> 
  transform(change =(x2020-x2010)/x2010)

# Not sure about the best visualization. Ideally a Shiny app that can tab across years to show the change in median rent.

leaf_sf_col(r1_rent_change_map,
            col_var = change)


```

There are clear differences in median rents between the northern and southern sections of the Route 1 corridor. Northern Chester and southern Bellwood had median rents above \$1,300 in 2020 --- \$1,314 and \$1,367, respectively. Median rents in the northern part of the corridor ranged from \$780 to \$1,059 -- hundreds of dollars less in all cases than the southern part of the corridor. Much of this difference in rental markets is no doubt a result of newer developments like Colony Village and The Jane at Moore's Lake. There have been few new multifamily developments in the northern section aside from Better Housing Coalition developments.

```{r acs-r1-rent1}

r1_rent_cpi <- r1_rent_cpi |> 
  mutate(neighborhood = case_when(
    NAMELSAD == "Census Tract 1003" ~ "Bensley",
    NAMELSAD == "Census Tract 1004.05" ~ "Bensley",
    NAMELSAD == "Census Tract 1004.04" ~ "Bellwood",
    NAMELSAD == "Census Tract 1004.06" ~ "Bellwood",
    NAMELSAD == "Census Tract 1004.07" ~ "Bellwood",
    NAMELSAD == "Census Tract 1004.09" ~ "Northern Chester",
  ))

gg_line_col_facet(r1_rent_cpi,
            x_var = year,
            y_var = dollars20,
            col_var = NAMELSAD,
            facet_var = neighborhood
            )


```

#### Multifamily Rents

CoStar's analytics allows for a broad snapshot of average rent data along Route 1 on a quarter-by-quarter basis. Average rent per apartment was on a steady decline from 2002 until the rental market began to recover in 2013. It was not until Q3 2014 that average rent began to experience consistent increases quarter over quarter from \$1,109 to \$1,214 in the most recent 2022 quarter. From 2017 to early 2020, the typical average rent increase was about \$3 but between Q1 2020 and Q2 2020 the average rent jumped \$15 from \$1,195 to \$1,210. It was during this timeframe that the pandemic began to impact housing markets, leading to significantly low rental vacancy rates as the eviction moratorium was put in place and increasing demand for housing away from the denser urban environment.

```{r costar-r1-rent}

# Use CoStar Multi-family properties search. Draw with polygon tool on map the outline of the Route One census tracts. Filter out properties from previous. Download inflation adjusted history for all years. 

costar_r1_rent <- read_csv("data/costar_r1_rent.csv") |> 
  clean_names() |> 
  mutate(period = case_when(
    period == "2022 Q3 QTD" ~ "2022 Q3",
    TRUE ~ period)) |> 
  mutate(period = yq(period)) |> 
  mutate(current_search = parse_number(current_search))

colnames(costar_r1_rent)[2] <- "rent"

costar_r1_rent_line <- gg_line(costar_r1_rent,
        x_var = period,
        y_var = rent) + 
  scale_x_date()

plotly::ggplotly(costar_r1_rent_line) |> 
  plotly_camera()


```

Average rent disaggregated by number of bedrooms shows fairly steady increases for all bedroom types, especially one- and two-bedroom apartments. Studios and three-bedroom apartments in the area have seen decreasing rents in recent years. In fact, average rent for three bedroom apartments are now \$56 cheaper (\$1,431) than they were at the start of 2012 (\$1,487). While we can only speculate as to reason for this, there has been a growing shift investor

```{r costar-r1-brent}

# Use CoStar Multi-family properties search. Draw with polygon tool on map the outline of the Route One census tracts. Filter out properties from previous. Download inflation adjusted history for all years. 


costar_r1_brent <- read_csv("data/costar_r1_brent.csv") |> 
  clean_names() |> 
  mutate(period = case_when(
    period == "2022 Q3 QTD" ~ "2022 Q3",
    TRUE ~ period)) |> 
  pivot_longer(
    cols = 2:5,
    names_to = "br",
    values_to = "rent"
  ) |> 
  mutate(br = case_when(
    br == "studio" ~ "Studio",
    br == "x1_bed" ~ "One bedroom",
    br == "x2_beds" ~ "Two bedroom",
    br == "x3_beds" ~ "Three bedroom"
  )) |> 
  mutate(period = yq(period)) |> 
  mutate(year = year(period)) |> 
  mutate(rent = parse_number(rent)) |> 
  filter(year > 2011)

r1_br_line<- gg_line_col_facet(costar_r1_brent,
        x_var = period,
        y_var = rent,
        col_var = br,
        facet_var = br) + 
  scale_x_date()

plotly::ggplotly(r1_br_line) |> 
  plotly_camera()

```

#### Single-family rents

Single-family rentals are not often listed via sources that provide consistent or reliable data. Facebook groups and other social networks often play a major role in advertising SFR, but the Central Virginia Regional Multiple Listing Service (CVR MLS) does pick up some of this market. Within the last year, there were 44 single-family homes rented through the MLS.

The average listed rent for one of these SFR was \$1,336 --- nearly \$100 less the average rent for a three-bedroom apartment. Single-family detached homes fetch higher rental prices versus condominiums or duplexes.

```{r mls-sfr}

# Use MLS Search RNT search. Filters: Status/Date Range: Active, Leased - 0-365; Map: draw outline for census tracts and export.

mls_r1_sfr <- read_csv("data/mls_r1_sfr.csv") |> 
  clean_names() |> 
  mutate(sales_date = mdy(sales_date)) |> 
  mutate(list_price = parse_number(list_price))

gg_point_col(mls_r1_sfr,
         x_var = sales_date,
         y_var = list_price,
         col_var = type)

mean(mls_r1_sfr$list_price, na.rm = TRUE)
```

The majority of SFR listed through MLS are three-bedroom homes (54 percent), while 18 percent are four-bedrooms or more --- an option that apartments do not often offer.

```{r mls-sfr1}

# Spread of bedrooms

mls_r1_sfr_br <- mls_r1_sfr |> 
  mutate(number_bedrooms = as.character(number_bedrooms)) |> 
  group_by(number_bedrooms) |> 
  summarise(count = n_distinct(address)) |> 
  transform(percent = count/sum(count))

gg_bar_col(mls_r1_sfr_br,
           x_var = number_bedrooms,
           y_var = percent,
           col_var = number_bedrooms)


```

### Conditions

Aging rental properties require substantial capital in order to maintain properties. Even properties built in the 1980s are approaching half a century in age. Along the Route 1 corridor, there are 1,812 rental homes that were built before 1980. The largest share of these homes are located in the Bensley area (Census Tracts 1003 and 1004.04).

```{r acs-yrblt}

b25036_vars <- load_variables(2020, "acs5") |> 
  filter(str_sub(name, end = 6) %in% "B25036")

b25036_raw <- get_acs(
  geography = "tract",
  county = "Chesterfield County",
  state = "VA",
  table = "B25036",
  year = 2020,
  survey = "acs5",
  cache_table = TRUE,
  geometry = TRUE
  )

b25036_r1_raw <- b25036_raw |> 
  subset(GEOID %in% r1_tracts$GEOID)

# Maybe keep all households in order to do percentages across entire housing stock? Or just rental housing stock?

b25036_vars_cleaned <- b25036_vars |> 
  separate(label, into = c("est", "total", "tenure", "yrblt"), sep = "!!") |> 
  select(variable = name, tenure, yrblt) |> 
  filter(tenure == "Renter occupied:") |> 
  drop_na() |> 
  mutate(tenure = "Renter")

b25036_r1 <- b25036_r1_raw |> 
  right_join(b25036_vars_cleaned, by = "variable")

b25036_r1_grp <- b25036_r1 |> 
  mutate(yrblt = case_when(
    yrblt == "Built 1939 or earlier" ~ "1939 or earlier",
    yrblt == "Built 1940 to 1949" ~ "1940 to 1959",
    yrblt == "Built 1950 to 1959" ~ "1940 to 1959",
    yrblt == "Built 1960 to 1969" ~ "1960 to 1979",
    yrblt == "Built 1970 to 1979" ~ "1960 to 1979",
    yrblt == "Built 1980 to 1989" ~ "1980 to 1999",
    yrblt == "Built 1990 to 1999" ~ "1980 to 1999",
    yrblt == "Built 2000 to 2009" ~ "2000 to 2009",
    yrblt == "Built 2010 to 2013" ~ "2010 or later",
    yrblt == "Built 2014 or later" ~ "2010 or later"
  )) |> 
  mutate(yrgrp = case_when(
    yrblt == "2000 to 2009" ~ yrblt,
    yrblt == "2010 or later" ~ yrblt,
    yrblt == "1980 to 1999" ~ yrblt,
    TRUE ~ "Pre-1980"
  )) |> 
  mutate(total_renters = sum(estimate))

# Total renter households is 3,376

b25036_r1_grp <- b25036_r1_grp |> 
  group_by(yrgrp, NAME) |> 
  summarise(estimate = sum(estimate)) |> 
  filter(yrgrp == "Pre-1980") |> 
  transform(percent = estimate/3376)

# Map showing concentration of pre-1980 rentals.

gg_sf_col(b25036_r1_grp,
          col_var = percent) +
  scale_fill_viridis_c(option = "magma")
  


```

Of the 23 CoStar-listed multifamily properties, 14 (61 percent) were constructed before 1980. There was no major multifamily development activity again until 2000 when Better Housing Coalition constructed Winchester Green Apartments' first phase. There has been little new multifamily development along the corridor during the last two decades --- activity has largely been spearheaded by Better Housing Coalition and Emerson Companies. With few newer developments, there are fewer options for those who are looking to rent along the corridor --- either having to choose a pre-1980s apartment, which are often in need of renovation, or among new luxury apartments that are often outside the budgets of existing renters along the corridor.

```{r costar-mf-3}

# Histogram or chart?

costar_r1_yrblt <- costar_r1_mf |> 
  select(year_built, property_address, building_status)

gg_histogram(costar_r1_yrblt,
             x_var = year_built)

```

CoStar's building rating system also provides a snapshot of condition of multifamily properties. Eighty-three percent of multifamily properties along the corridor are three-star or less --- meaning that in terms of quality and amenities that these buildings are relatively average or simply functional. In comparison, four-star properties are "are constructed with higher end finishes and specifications, providing desirable amenities to
residents and designed/built to competitive and contemporary standards," while five-star properties are "the luxury end of multi-family buildings defined by finishes, amenities, the overall interior/exterior design and the highest level of specifications for its style..."

```{r costar-mf-3}

costar_r1_stars <- costar_r1_mf |> 
  select(star_rating, property_address, building_status) |> 
  mutate(star_rating = as.character(star_rating)) |> 
  group_by(star_rating) |> 
  summarise(count = n_distinct(property_address))

  
r1_star_bar <- gg_bar_col(costar_r1_stars,
           x_var = star_rating,
           y_var = count,
           col_var = star_rating
           )

plotly::ggplotly(r1_star_bar) |> 
  plotly_camera()

```

-   MHC ratings from 2016 MHCCV study

## Proximity to amenities

### Public transportation

The Route 1 corridor has recently benefited from the 

```{r r1-stops}

costar_r1_map <- costar_r1_mf |>
  st_as_sf(coords = c("longitude", "latitude"),
           crs = 4326)

va_crs <- suggest_top_crs(costar_r1_map, units = "m",
                          inherit_gcs = FALSE)

r1_prj <- st_transform(costar_r1_map, va_crs)

r1_tracts_map <- tracts("VA", "Chesterfield", year = 2021) |>
  st_transform(va_crs)

costar_r1_tracts <- st_filter(r1_tracts_map, r1_prj)

stops <- st_read("data/grtc/Stops_May2022.shp") |> 
  st_transform(va_crs)

r1_network_buffer <- mb_isochrone(r1_prj, profile = "walking",
                                       distance = 805)
library(nngeo)

nearest_stops <- st_nn(r1_prj, stops, k = 20)

mapview(stops) + 
  mapview(costar_r1_tracts)

```

```{r r1-mf-stops}

costar_stop_walking <- imap(nearest_stops, ~{
  
  property <- r1_prj[.y,]
  
  nearby_stops <- stops[.x, ]

  mb_matrix(origins = property,
            destinations = nearby_stops,
            profile = "walking",
            output = "distance")}) |> 
  reduce(rbind) |> 
  apply(1, min) |> 
  magrittr::divide_by(1609.34)

r1_trans_map <- costar_r1_map |>
   mutate(
     walk_to_transit = costar_stop_walking,
     transit_nearby = ifelse(walk_to_transit <= 0.5, 1, 0)
   )


stops_shp <- readOGR(dsn = ".",
                     layer = "data/Stops_May2022",
                     verbose = FALSE)

stops_latlon <- spTransform(stops_shp, CRS("+proj=longlat +datum=WGS84"))


pal <- colorFactor(c("red", "green"), domain = c(0, 1))

# How can I add a stops layer to show the location of stops to this. There is some issue with the stops data frame when adding in leaflet. 

leaflet(r1_trans_map) |>
   addTiles() |>
   addCircleMarkers(
     radius = 5,
     color = ~pal(transit_nearby),
     stroke = FALSE, fillOpacity = 0.5
    )

mapview(r1_trans_map, zcol = "transit_nearby") +
  mapview(stops)

mean(r1_trans_map$walk_to_transit)

```

Compare location of new multifamily development versus older multifamily relative to access to public transportation

-   Within or outside of 1/2 mile buffer from GRTC stop

### Schools

```{r r1-school-map}

schools <- st_read("https://services3.arcgis.com/TsynfzBSE6sXfoLq/ArcGIS/rest/services/Administrative/FeatureServer/15/query?outFields=*&where=1%3D1&f=geojson") |> 
  st_transform(va_crs)

nearest_schools <- st_nn(r1_prj, schools, k = 20)


mapview(schools) +
  mapview(costar_r1_tracts)

```


```{r r1-schools}
r1_school_driving <- imap(nearest_schools, ~{
  
  property <- r1_prj[.y,]

  nearby_schools <- schools[.x, ]

  mb_matrix(origins = property,
            destinations = nearby_schools,
            profile = "driving",
            output = "duration")
}) |> 
  reduce(rbind) |> 
  apply(1, min)

r1_schools_map <- r1_trans_map |>
  mutate(
    minutes_to_school = r1_school_driving,
    school_nearby = ifelse(minutes_to_school <= 30, 1, 0))

```

Compare location of new multifamily development versus older multifamily relative to access to public schools

-   Within or outside of 30 minute drive time

### Grocery stores

Compare location of new multifamily development versus older multifamily relative to access to grocery stores

-   Within or outside of 1/2 mile buffer from grocery store

### Healthcare

Compare location of new multifamily development versus older multifamily relative to access to hospitals and urgent care facilities

-   Within or outside of 15 minute drive time

### Job centers

TBD, waiting for feedback from county staff

## Public and private investments

-   Check CoStar for newest builds/renovations/transactions (commercial properties, too)
-   Ask Jessica for what county has had a hand in

Route 10, Bermuda Estates, Falling Creek Rehab, Multifamily Overlay District, R.O.A.R.

### Case study

-   Choose an interesting recent residential development
-   Add a photo
-   Describe development and market it serves
-   Rent/sale prices

## Single-family rental versus purchase

-   Single-family rental prices from MLS records
-   Median home sales price from past 365 days
-   Calculate income needed to afford SFR and homeownership
-   Use same methodology from Part 1b-4 for homeownership affordability
