# Eastern Hull Street Road corridor analysis {#part-2b-1}

```{r setup}

library(tidycensus)
library(tidyverse)
library(janitor)
library(leaflet)
library(sf)
library(sp)
library(tigris)
library(mapview)
library(simplevis)
library(readxl)
library(ggplot2)
library(lubridate)
library(viridis)
library(tidygeocoder)
library(knitr)
library(kableExtra)
library(crsuggest)
options(tigris_use_cache = TRUE)
remotes::install_github("walkerke/mapboxapi")
library(mapboxapi)
library(rgdal)
library(optiRum)
library(ggrepel)

```

This section describes the Eastern Hull Street Road corridor in terms of its rental housing market, proximity of rental homes to amenities, recent major investments, and homeownership relative to the price of single-family rentals.

## The Hull Street Corridor

For the purposes of this analysis, this corridor is defined as the Census tracts 1002.05, 1002.06, 1002.08, 1002.09, 1002.10, and 1002.11. These tracts include many of the neighborhoods directly adjacent to Hull Street Road from Chippenham Parkway to Courthouse Road.

It should be noted that portions of Census tracts 1002.09 and 1002.10 span from Hull Street to Midlothian Turnpike. This results in the inclusion of rental properties that are not technically tied to the Hull Street corridor. Census tract analysis allows for consistency between the analyses conducted here and the previously in [Chapter 15](#part-2b-1) with the Route 1 corridor.

Unlike Route 1, Hull Street (Route 360) does not have the storied interstate tourist history that has since evaporated with the development of the interstate highway system. Along Hull Street towards the City of Richmond, there was significant commercial development that fronted Hull Street, most of which has experienced decline and disinvestment.

But much of the development along Hull Street in Chesterfield County today is much less dense. The corridor in the county remains car-centric with no multifamily properties fronting the corridor and no manufactured home communities. Most development along the corridor is concentrated towards the eastern end at the city-county border, while development fronting the corridor becomes more rare west of Pocoshock Creek and the new Manchester Middle School.

Opportunities exist for multifamily development along the Hull Street corridor, but there has been none in recent years and there are none planned for the foreseeable future. 

```{r hsr-base-map, fig.cap="Eastern Hull Street Road corridor", message=FALSE}

hsr_tracts_list <- c("100205", "100206", "100208", "100209", "100210", "100211")

hsr_tracts <- tracts("VA", "Chesterfield",
                    progress_bar = FALSE) |>  
  filter(TRACTCE %in% hsr_tracts_list) |> 
  st_transform(4326)

# Run the mapview and savewidget functions before building the book
# Once the widget is saved, activate the knitr function to embed the map

mapview(hsr_tracts,
          alpha = 0.5,
          col.regions = "cornflowerblue",
          legend = F,
          popup = F,
          label = "NAMELSAD",
          layer.name = "Eastern Hull Street Road corridor census tracts")
#
# htmlwidgets::saveWidget(hsr_map@map, "maps/hsr_map.html")

# knitr::include_url("maps/hsr_map.html", height = "500px")

```

## Rental housing supply

### Number of rental homes

Despite the low number of multifamily properties in the Hull Street corridor, the number of renters --- like all areas of the county --- has experienced significant increases. From 1,054 renter households in 2010 to 2,388 in 2020, renter households have more than doubled in a decade (a 126 percent increase). Much of this growth occurred between 2018 and 2020, when renter households increased 78 percent from 1,337.

```{r hsr-rhh}

years <- 2010:2020

b25003_vars <- load_variables(2020, "acs5") |> 
  filter(str_sub(name, end = 7) %in% "B25003_")

b25003_raw <- map_dfr(years, function(yr){
  b25003_pull <- get_acs(
    geography = "tract",
    county = "Chesterfield County",
    state = "VA",
    table = "B25003",
    year = yr,
    survey = "acs5",
    cache_table = TRUE,
    geometry = TRUE
  ) |> 
    mutate(year = yr)
  })

b25003_hsr_raw <- b25003_raw |> 
  subset(GEOID %in% hsr_tracts$GEOID)

b25003_vars_cleaned <- b25003_vars |> 
  separate(label, into = c("est", "total", "tenure"), sep = "!!") |> 
  select(variable = name, tenure) |> 
  drop_na() |> 
  mutate(tenure = case_when(
    tenure == "Owner occupied" ~ "Homeowner",
    tenure == "Renter occupied" ~ "Renter"
  )) |> 
  filter(tenure == "Renter")

b25003_hsr <- b25003_hsr_raw |> 
  right_join(b25003_vars_cleaned, by = "variable") |> 
  select(NAME, fips = GEOID, year, tenure, estimate, moe)

b25003_r360 <- b25003_hsr |> 
  group_by(year) |> 
  summarise(estimate = sum(estimate))
  
#Create line graph showing change in renter households across time.
hsr_renters_plot <- gg_line(
  b25003_r360,
  x_var = year,
  y_var = estimate,
  y_breaks_n = 5,
  title = "Renter households",
  x_title = element_blank())

plotly::ggplotly(hsr_renters_plot) %>% 
  plotly_camera()

  
```
Based on a survey of CoStar listed multifamily properties in the Hull Street corridor study area, there are only 1,198 apartments --- accounting for only half of renter households in the area. This begs the question as to where many renter households find housing when there are so few options in a typical apartment setting. A large share of these apartments are oriented towards the Midlothian Turnpike corridor at the recently developed Element at Stonebridge apartments --- 400 mid-rise apartments built in 2016.

```{r costar-hsr-1, fig.cap= "Total number of multifamily units by property status"}

# Load .csv of CoStar properties. Data was collected by hand drawing boundaries for the necessary census tracts in the CoStar platform. No filters were placed on the search aside from multi-family property. Need to filter out Sandler - it's a single condo.


costar_hsr_mf <- read_csv("data/costar_hsr_mf.csv") |> 
  clean_names() |> 
  filter(property_address != "7626 Sandler Dr") |> 
  mutate(corridor = case_when(
    property_address == "301 Karl Linn Dr" ~ "Midlothian",
    property_address == "3005 Lancers Blvd" ~ "Hull Street",
    property_address == "6300 Pewter Ave" ~ "Hull Street",
    property_address == "6925 Starview Ct" ~ "Midlothian",
    property_address == "7660 Van Hoy Dr" ~ "Hull Street",
    property_address == "3101 Danzler Cir" ~ "Hull Street"
  )) |> 
  mutate(property_name = case_when(
    property_address == "3101 Danzler Cir" ~ "Danzler Place Townhomes",
    TRUE ~ property_name
  ))

costar_hsr_mf |> 
  arrange(desc(number_of_units)) |> 
  select(5, 33, 7) |> 
  kable(caption = "Multifamily properties located in Hull Street corridor study area",
        align = "llr",
        col.names = c("Name", "Corridor", "Estimated apartment homes")) |> 
  kableExtra::kable_styling(
    bootstrap_options = c("condensed", "hover", "striped"))
```
No multifamily properties actually front the Eastern Hull Street corridor. The four multifamily properties in closest proximity to the corridor are more closely associated with the Walmsley Boulevard (State Route 647) corridor.

```{r costar-hsr-2, fig.cap="Map of multifamily properties in Eastern Hull Street study area"}

hsr_map <- costar_hsr_mf |> 
  st_as_sf(coords = c("longitude", "latitude"), crs=4326, remove=FALSE)

leaflet(hsr_map) |> 
  addTiles() |> 
  addCircleMarkers(
    radius = 3) |> 
  addPolygons(data=hsr_tracts, weight = 2, fillColor = NULL)


```
With no manufactured home communities in the area, single-family home rentals (SFRs) are the option left for many Hull Street corridor residents. An estimated 900 SFRs are located throughout the study area - 14 percent of all single-family detached homes.

Higher concentrations of SFR are located in Census tracts 1002.10 and 1002.11; tracts that adjacent to the Chippenham Parkway and along the city-county border. These tracts have an estimated 21 percent and 20 percent of single-family detached homes being utilized as SFR, respectively. 

```{r sfr-hsr}

# Choropleth map of percent of SFR by census tract. Perhaps include a chart as well. Might be interesting to compare to change in homeownership rate for each.

sfr_hsr <- ch_sfh |> 
  st_join(left = FALSE, hsr_tracts["NAMELSAD"]) |> 
  select(address, year_built, improvement, ownership, owner_zip, owner_state, NAMELSAD, geometry) |> 
  filter(improvement > 0)

sfr_hsr_summary <- sfr_hsr |> 
  group_by(NAMELSAD, ownership) |> 
  summarise(count = n_distinct(address)) |> 
  mutate(total_sfh = case_when(
    NAMELSAD == NAMELSAD ~ sum(count))
  ) |> 
  mutate(percent_sfh = count/total_sfh) |> 
  filter(ownership == "Renter-occupied SFH") |> 
  st_drop_geometry()

sfr_hsr_tracts <- hsr_tracts |> 
  left_join(sfr_hsr_summary, by = "NAMELSAD")

leaf_sf_col(sfr_hsr_tracts,
            col_var = percent_sfh)

sum(sfr_hsr_summary$count)

sum(sfr_hsr_tracts$total_sfh)
```

### Rental prices

Rent prices, as estimated by the American Community Survey (ACS), show very little change in rental prices along the Hull Street corridor, except in Census tract 1002.08. This Census tract encompasses a large portion of the Rockwood neighborhood, which has no recorded multifamily properties. Nonetheless, the estimated median gross rent in this area has increased by 64 percent from 2010 to 2020 --- from \$1,231 to \$2,021. Margins of error for this tract remain high in 2019 and 2020 - roughly +/- \$1,000 in both years --- likely due to small sample sizes.

Other Census tracts in the area have more consistent rent increases - particularly Census tracts 1002.06, 1002.09, and 1002.10. Census tract 1002.11 does not account for a median gross rent until 2020 (\$980), which is strange given that CoStar recognizes three multifamily properties in this Census tract, all of which were built pre-1980.

```{r acs-hsr-rent}

# Check MOE for the census tract above

years <- 2010:2020

b25064_vars <- load_variables(2020, "acs5") |> 
  filter(str_sub(name, end =6) %in% "B25064")

b25064_raw <- map_dfr(years, function(yr){
  b25064_pull <- get_acs(
    geography = "tract",
    county = "Chesterfield County",
    state = "VA",
    table = "B25064",
    year = yr,
    survey = "acs5",
    cache_table = TRUE,
    geometry = TRUE
  ) %>%
    mutate(year = yr)
  })

b25064_hsr_raw <- b25064_raw |> 
  subset(GEOID %in% hsr_tracts$GEOID)

cpi <- read_excel("data/CPI_U_RS.xlsx")
cpi <- cpi |>  
  rename(year = Year,
         priceindex = Index) |> 
  transform(year = as.numeric(year))

hsr_rent_cpi <- b25064_hsr_raw |> 
    left_join(cpi, by = 'year') |> 
  transform(dollars20 = ((381.2/priceindex)*estimate)) |> 
  select(GEOID, NAME, year, dollars20, moe) |> 
  separate(NAME, into = c("NAMELSAD", "county", "state"), sep = ",") |> 
  transform(year_char = as.character(year))

hsr_rent_change <- hsr_rent_cpi |> 
  select(NAMELSAD, year, dollars20) |> 
  st_drop_geometry() |> 
  pivot_wider(
    names_from = year,
    values_from = dollars20
  )

hsr_rent_change_map <- hsr_tracts |>
  left_join(hsr_rent_change, by = "NAMELSAD") |> 
  clean_names() |> 
  select(tract = namelsad, x2010:x2020) |> 
  transform(change =(x2020-x2010)/x2010)

# Not sure about the best visualization. Ideally a Shiny app that can tab across years to show the change in median rent.

leaf_sf_col(hsr_rent_change_map,
            col_var = change)


```



```{r acs-hsr-rent1}

hsr_rent_cpi$label <- NA
hsr_rent_cpi$label[which(hsr_rent_cpi$year == max(hsr_rent_cpi$year))] <- 
  hsr_rent_cpi$dollars20[which(hsr_rent_cpi$year == max(hsr_rent_cpi$year))] 


ggplot(hsr_rent_cpi, aes(x = year, y = dollars20)) +
  geom_line(aes(color = NAMELSAD)) +
  geom_point(aes(color = NAMELSAD)) +
  scale_y_continuous(labels = scales::dollar_format()) +
  ggplot2::scale_x_continuous(limits = c(2009.5, 2020.5),
                              breaks = seq(2010, 2020, 1)) +
  labs(title = "Renter income in the Eastern Hull Street corridor study area",
       x = NULL, y = "Median household income (2020 dollars)") +
  scale_color_discrete(name = "Legend")



```

Focusing on the areas adjacent to the Hull Street corridor, CoStar's data show an average rent across the four existing properties. Changes in average rent have been slight over the past decade, but have been on a steady climb --- reaching a decade max at the start of 2022 with an average rent of \$1,140. Over the course of the pandemic, rents in the area have seen upticks much like other parts of the county, including the Route 1 corridor. From Q1 2020 to Q3 2022, average rent saw a 2 percent increase from \$1,088 to \$1,110. But in 2022 so far, average rent has been in decline from its decade max --- potentially signaling some cooling of the hot pandemic housing market.

```{r costar-hsr-rent}

# Use CoStar Multi-family properties search. Draw with polygon tool on map the outline of the Route One census tracts. Filter out properties from previous. Download inflation adjusted history for all years. I have two versions of this data pull - one that excludes the developments near Midlo Turnpike and one that includes them --- including Midlo results in some erratic rent changes.

# I advise keeping Element and Cloverleaf properties out of the analysis for those reasons.

costar_hsr_rent <- read_csv("data/costar_hsr_rent_nomidlo.csv") |> 
  clean_names() |> 
  mutate(period = case_when(
    period == "2022 Q3 QTD" ~ "2022 Q3",
    TRUE ~ period)) |> 
  mutate(period = yq(period)) |> 
  mutate(current_search = parse_number(current_search)) |> 
  mutate(year = year(period)) |>
  filter(year > 2011)

colnames(costar_hsr_rent)[2] <- "rent"

costar_hsr_rent_line <- gg_line(costar_hsr_rent,
        x_var = period,
        y_var = rent) + 
  scale_x_date()

plotly::ggplotly(costar_hsr_rent_line) |> 
  plotly_camera()


```
Breaking out average rent by bedrooms shows a similar pattern among all bedroom sizes, except for studio apartments, in which there are none listed in the study area. 

```{r costar-hsr-brent}

# Use CoStar Multi-family properties search. Draw with polygon tool on map the outline of the Route One census tracts. Filter out properties from previous. Download inflation adjusted history for all years. 


costar_hsr_brent <- read_csv("data/costar_hsr_brent_nomidlo.csv") |> 
  clean_names() |> 
  mutate(period = case_when(
    period == "2022 Q3 QTD" ~ "2022 Q3",
    TRUE ~ period)) |> 
  pivot_longer(
    cols = 2:5,
    names_to = "br",
    values_to = "rent"
  ) |> 
  mutate(br = case_when(
    br == "studio" ~ "Studio",
    br == "x1_bed" ~ "One bedroom",
    br == "x2_beds" ~ "Two bedroom",
    br == "x3_beds" ~ "Three bedroom"
  )) |> 
  mutate(period = yq(period)) |> 
  mutate(year = year(period)) |> 
  mutate(rent = parse_number(rent)) |> 
  filter(year > 2011) |> 
  filter(br != "Studio")

hsr_br_line<- gg_line_col_facet(costar_hsr_brent,
        x_var = period,
        y_var = rent,
        col_var = br,
        facet_var = br) + 
  scale_x_date()

plotly::ggplotly(hsr_br_line) |> 
  plotly_camera()

```

Very little rental price data exists for SFRs in the study area. Across the past four years, there have only been 25 listed SFRs. Thirteen of these 25 SFRs are single-family detached homes, while the remaining are largely condominums or townhomes (11). List prices have ranged from \$900 for a duplex in 2019 to a /$1,809 for a single-family detached home in 2022. Generally, single-family detached homes have commanded rents above \$1,400 and often offer three-bedrooms and two baths.These rents are on par with recent two-bedroom apartments rents in the area --- often making a decision between an apartment or single-family detached home on the rental market easy for a family of three or more.

```{r mls-sfr}

# Use MLS Search RNT search. Filters: Status/Date Range: Active, Leased - 4 years; Map: draw outline for census tracts and export. From July 2022.

mls_hsr_sfr <- read_csv("data/mls_hsr_sfr.csv") |> 
  clean_names() |> 
  mutate(sales_date = mdy(sales_date)) |> 
  mutate(list_price = parse_number(list_price))

gg_point_col(mls_hsr_sfr,
         x_var = sales_date,
         y_var = list_price,
         col_var = type)

mean(mls_hsr_sfr$list_price, na.rm = TRUE)
```

### Conditions

Much of the rental housing stock (62 percent) in the Eastern Hull Street corridor was built after 1980 --- 1,472 rental homes. The largest share of occupied rental housing was built in the latter half of the 20th century (1960 to 1999). These buildings are anywhere from 20 to 50 years old and are most likely in need of modest to major upgrades.

```{r acs-yrblt}

b25036_vars <- load_variables(2020, "acs5") |> 
  filter(str_sub(name, end = 6) %in% "B25036")

b25036_raw <- get_acs(
  geography = "tract",
  county = "Chesterfield County",
  state = "VA",
  table = "B25036",
  year = 2020,
  survey = "acs5",
  cache_table = TRUE,
  geometry = TRUE
  )

b25036_hsr_raw <- b25036_raw |> 
  subset(GEOID %in% hsr_tracts$GEOID)

# Maybe keep all households in order to do percentages across entire housing stock? Or just rental housing stock?

b25036_vars_cleaned <- b25036_vars |> 
  separate(label, into = c("est", "total", "tenure", "yrblt"), sep = "!!") |> 
  select(variable = name, tenure, yrblt) |> 
  filter(tenure == "Renter occupied:") |> 
  drop_na() |> 
  mutate(tenure = "Renter")

b25036_hsr <- b25036_hsr_raw |> 
  right_join(b25036_vars_cleaned, by = "variable")

b25036_hsr_grp <- b25036_hsr |> 
  mutate(yrblt = case_when(
    yrblt == "Built 1939 or earlier" ~ "1939 or earlier",
    yrblt == "Built 1940 to 1949" ~ "1940 to 1959",
    yrblt == "Built 1950 to 1959" ~ "1940 to 1959",
    yrblt == "Built 1960 to 1969" ~ "1960 to 1979",
    yrblt == "Built 1970 to 1979" ~ "1960 to 1979",
    yrblt == "Built 1980 to 1989" ~ "1980 to 1999",
    yrblt == "Built 1990 to 1999" ~ "1980 to 1999",
    yrblt == "Built 2000 to 2009" ~ "2000 to 2009",
    yrblt == "Built 2010 to 2013" ~ "2010 or later",
    yrblt == "Built 2014 or later" ~ "2010 or later"
  )) |> 
  mutate(yrgrp = case_when(
    yrblt == "2000 to 2009" ~ yrblt,
    yrblt == "2010 or later" ~ yrblt,
    yrblt == "1980 to 1999" ~ yrblt,
    TRUE ~ "Pre-1980"
  )) |> 
  mutate(total_renters = sum(estimate))


b25036_hsr_summary <- b25036_hsr_grp |> 
  group_by(yrblt) |> 
  summarise(estimate = sum(estimate)) |> 
  st_drop_geometry()

# Total renter households is 2,388

b25036_hsr_1980 <- b25036_hsr_grp |> 
  group_by(yrgrp, NAME) |> 
  summarise(estimate = sum(estimate)) |> 
  filter(yrgrp == "Pre-1980") |> 
  transform(percent = estimate/2388)

# Map showing concentration of pre-1980 rentals.

gg_sf_col(b25036_hsr_1980,
          col_var = percent) +
  scale_fill_viridis_c(option = "magma")
  
```

Multifamily property construction has been rare in the corridor for much of the last half century. Only one development was completed in the 2000s --- Element at Stonebridge. This development was the result of the Cloverleaf Mall redevelopment and is oriented towards Midlothian Turnpike, rather than the Eastern Hull Street corridor. 

More relevant to the corridor is Cross Creek Apartments, the most recent development in close proximity to the Eastern Hull Street. This 225 unit garden style apartment property was completed in 1991 --- 31 years ago --- and was the first multifamily development along the Eastern Hull Street corridor since 1978. 
```{r costar-hsr-3}

# Histogram or chart?

costar_hsr_yrblt <- costar_hsr_mf |> 
  select(property_name, corridor, year_built)

costar_hsr_yrblt |> 
  arrange(desc(year_built)) |> 
  kable(caption = "Multifamily properties located in Hull Street corridor study area by year built",
        align = "llr",
        col.names = c("Name", "Corridor", "Year Built")) |> 
  kableExtra::kable_styling(
    bootstrap_options = c("condensed", "hover", "striped"))


```

- CoStar rating

```{r costar-hsr-4}

costar_hsr_stars <- costar_hsr_mf |> 
  select(star_rating, property_address, building_status) |> 
  mutate(star_rating = as.character(star_rating)) |> 
  group_by(star_rating) |> 
  summarise(count = n_distinct(property_address))

  
hsr_star_bar <- gg_bar_col(costar_hsr_stars,
           x_var = star_rating,
           y_var = count,
           col_var = star_rating
           )

plotly::ggplotly(hsr_star_bar) |> 
  plotly_camera()

```

## Proximity to amenities

### Public transportation

Compare location of new multifamily development versus older multifamily relative to access to public transportation

- Within or outside of 1/2 mile buffer from GRTC stop

```{r r1-stops}

costar_hsr_map <- costar_hsr_mf |>
  st_as_sf(coords = c("longitude", "latitude"),
           crs = 4326)

va_crs <- suggest_top_crs(costar_hsr_map, units = "m",
                          inherit_gcs = FALSE)

hsr_prj <- st_transform(costar_hsr_map, va_crs)

hsr_tracts_map <- tracts("VA", "Chesterfield", year = 2021) |>
  st_transform(va_crs)

costar_hsr_tracts <- st_filter(hsr_tracts_map, hsr_prj)

stops <- st_read("data/grtc/Stops_May2022.shp") |> 
  st_transform(va_crs)

hsr_network_buffer <- mb_isochrone(hsr_prj, profile = "walking",
                                       distance = 805)
library(nngeo)

nearest_stops <- st_nn(hsr_prj, stops, k = 20)

mapview(stops) + 
  mapview(costar_hsr_tracts)

```

```{r r1-mf-stops}

costar_stop_walking <- imap(nearest_stops, ~{
  
  property <- hsr_prj[.y,]
  
  nearby_stops <- stops[.x, ]

  mb_matrix(origins = property,
            destinations = nearby_stops,
            profile = "walking",
            output = "distance")}) |> 
  reduce(rbind) |> 
  apply(1, min) |> 
  magrittr::divide_by(1609.34)

hsr_trans_map <- costar_hsr_map |>
   mutate(
     walk_to_transit = costar_stop_walking,
     transit_nearby = ifelse(walk_to_transit <= 0.5, 1, 0)
   )

pal <- colorFactor(c("red", "green"), domain = c(0, 1))

# How can I add a stops layer to show the location of stops in relation to properties. There is some issue with the stops data frame when adding in leaflet. 

leaflet(hsr_trans_map) |>
   addTiles() |>
   addCircleMarkers(
     radius = 5,
     color = ~pal(transit_nearby),
     stroke = FALSE, fillOpacity = 0.5
    ) 

mapview(hsr_trans_map, zcol = "transit_nearby") +
  mapview(stops)

mean(hsr_trans_map$walk_to_transit)

```
```{r r1-school-map}

schools <- st_read("https://services3.arcgis.com/TsynfzBSE6sXfoLq/ArcGIS/rest/services/Administrative/FeatureServer/15/query?outFields=*&where=1%3D1&f=geojson") |> 
  st_transform(va_crs)

nearest_schools <- st_nn(hsr_prj, schools, k = 20)

mapview(schools) +
  mapview(costar_hsr_tracts)

```

```{r r1-schools}

# I think I would want to understand the number of schools within a distance of certain properties and what schools those are. 

hsr_school_driving <- imap(nearest_schools, ~{
  
  property <- hsr_prj[.y,]

  nearby_schools <- schools[.x, ]

  mb_matrix(origins = property,
            destinations = nearby_schools,
            profile = "driving",
            output = "duration")
}) |> 
  reduce(rbind) |> 
  apply(1, min)

hsr_schools_map <- hsr_trans_map |>
  mutate(
    minutes_to_school = hsr_school_driving,
    school_nearby = ifelse(minutes_to_school <= 30, 1, 0))

```

Compare location of new multifamily development versus older multifamily relative to access to public schools

-   Within or outside of 30 minute drive time

### Grocery stores

```{r r1-grocery}
# Grocery stores:
# We'll use SNAP retailers from the USDA as a proxy here
# This is not perfect so some filtering may be necessary
# as it picks up on convenience stores as well
snap <- read_csv("https://opendata.arcgis.com/datasets/e9cc76a48ccb45628181ece7b2deb56d_0.csv") %>%
  filter(State == "VA") %>%
  st_as_sf(coords = c("X", "Y"), crs = 4326) %>%
  st_transform(va_crs)

nearest_snap <- st_nn(hsr_prj, snap, k = 20)

hsr_grocery <- imap(nearest_snap, ~{

  property <- hsr_prj[.y,]

  nearby_snap <- snap[.x, ]

  mb_matrix(origins = property,
            destinations = nearby_snap,
            profile = "walking",
            output = "distance")
}) %>%
  reduce(rbind) %>%
  apply(1, min) %>%
  magrittr::divide_by(1609.34)

hsr_grocery_map <- hsr_schools_map |>
  mutate(
    miles_to_snap = hsr_grocery,
    snap_nearby = ifelse(miles_to_snap <= 0.5, 1, 0))

pal <- colorFactor(c("red", "blue"), domain = c(0,1))

leaflet(hsr_grocery_map) |>
  addTiles() |>
  addCircleMarkers(
    radius = 5,
    color = ~pal(snap_nearby),
    stroke = FALSE, fillOpacity = 0.5
   )


```

Compare location of new multifamily development versus older multifamily relative to access to grocery stores

-   Within or outside of 1/2 mile buffer from grocery store

### Healthcare

```{r medical}
# Hospital / urgent care
# I'm not filtering by type but you might consider doing so
# There are some other datasets in HIFLD on care centers that may 
# be relevant as well.
hospitals <- st_read("https://services1.arcgis.com/Hp6G80Pky0om7QvQ/arcgis/rest/services/Hospitals_1/FeatureServer/0/query?outFields=*&where=1%3D1&f=geojson") %>%
  filter(STATE == "VA") %>%
  st_transform(va_crs)
# 
nearest_hospitals <- st_nn(hsr_prj, hospitals, k = 20)
# 
costar_hospital_driving <- imap(nearest_hospitals, ~{
  # Get the specific property (the list index)
  property <- hsr_prj[.y,]
#   
  # Get the nearby hospitals (the indices in the list element)
  nearby_hospitals <- hospitals[.x, ]
#   
  # Calculate the matrix (drive-times here instead)
  # Minutes are returned by default
  mb_matrix(origins = property,
            destinations = nearby_hospitals,
            profile = "driving",
            output = "duration")
}) %>%
  reduce(rbind) %>%
  apply(1, min)

hsr_health_map <- hsr_grocery_map |>
  mutate(
    minutes_to_hospital = costar_hospital_driving,
    hospital_nearby = ifelse(minutes_to_hospital <= 15, 1, 0))

```
Compare location of new multifamily development versus older multifamily relative to access to hospitals and urgent care facilities

- Within or outside of 15 minute drive time

### Job centers

TBD, waiting for feedback from county staff

## Public and private investments

The development patterns of the Eastern Hull Street corridor have largely been 

- Check CoStar for newest builds/renovations/transactions (commercial properties, too)
- Ask Jessica for what county has had a hand in

This is recent but just to the west of the corridor area:
https://richmondbizsense.com/2022/06/02/chesterfield-oks-plan-to-redevelop-rockwood-square-shopping-center/

New Manchester Middle School - primary investment in the area in recent years - county investment in school - family growth - https://www.nbc12.com/2020/08/25/chesterfield-holds-virtual-ribbon-cutting-new-school/

New townhome development near Elkhardt?

### Case study

- Choose an interesting recent residential development
- Add a photo
- Describe development and market it serves
- Rent/sale prices

- townhome or existing - Cross Creek or Rollingwood?

## Single-family rental versus purchase

- Single-family rental prices from MLS records
- Median home sales price from past 365 days
- Calculate income needed to afford SFR and homeownership
- Use same methodology from Part 1b-4 for homeownership affordability
