# Rental market summary {#part-2a-1}

This section summarizes the existing stock of rental units within the county and provides an analysis of recent price and production trends in the rental market.

```{r}
#| label: setup

library(tidyverse)
library(tidycensus)
library(glue)
library(esquisse)
library(plotly)
library(scales)
library(formattable)
library(simplevis)
library(lubridate)
library(patchwork)
library(knitr)
library(zoo)
library(janitor)
library(qs)
library(readxl)
library(sf)
library(stringdist)
library(leaflet)
library(htmlwidgets)
library(pals)
library(mapboxapi)
library(ggspatial)


```

## Rental housing stock

A growing population, housing affordability challenges, and [changing housing preferences](https://www.freddiemac.com/research/consumer-research/20180404-preference-for-renting) have all contributed to the expansion of rental housing throughout Chesterfield County. This strong demand for rental housing on a spectrum---from high-end apartments for working professionals to single room occupancy units.

### Total renters

Over the past decade, the number of renter households in the county has steadily increased. From 2010 to 2019, the number of renters in the county grew from 24,312 to 30,273---a 24 percent increase. 

From 2019 to 2020, there was a two percent decline in the renting population. But this decrease of 633 renters may be attributed to [the collection challenges of the 2020 Census count](https://apnews.com/article/lifestyle-health-coronavirus-pandemic-census-2020-government-and-politics-9aebdfb9c1b834352676d4b95f2d9ea0) and [historical undercounting of renters](https://censuscounts.org/whats-at-stake/will-you-count-renters-in-the-2020-census/). As rising interest rates and home sales prices persist, the number of renters in the county can be expected to rise as well---further contributing to the need for a diverse mix of rental housing opportunities.

```{r}
#| label: fig-tenure
#| fig-cap: "Number of renter households in Chesterfield County"

# Pull number of renter households across time from ACS Table B25003.

years <- 2010:2020

b25003_vars <- load_variables(2010, "acs5") |> 
  filter(str_sub(name, end = 7) %in% "B25003_")

b25003_raw <- map_dfr(years, function(yr){
  b25003_pull <- get_acs(
    geography = "county",
    county = "Chesterfield County",
    state = "VA",
    table = "B25003",
    year = yr,
    survey = "acs5",
    cache_table = TRUE
  ) %>%
    mutate(year = yr)
  })

b25003_vars_cleaned <- b25003_vars |> 
  separate(label, into = c("est", "total", "tenure"), sep = "!!") |> 
  select(variable = name, tenure) |> 
  drop_na() |> 
  mutate(tenure = case_when(
    tenure == "Owner occupied" ~ "Homeowner",
    tenure == "Renter occupied" ~ "Renter"
  )) |> 
  filter(tenure == "Renter")

b25003_data <- b25003_raw |> 
  right_join(b25003_vars_cleaned, by = "variable") |> 
  select(NAME, fips = GEOID, year, tenure, estimate, moe)
  
# Create line graph showing change in renter households across time.

renters_plot <- gg_line(
  b25003_data,
  x_var = year,
  y_var = estimate,
  y_breaks_n = 5,
  y_zero = TRUE,
  title = "Renter households",
  x_title = element_blank())

renters_plot

# plotly::ggplotly(renters_plot) %>% 
#   plotly_camera()

```

### Rental housing type

The rise in renters in Chesterfield County has not led to a commensurate rise in diverse housing options for renters. Over a third (37 percent) of rental housing in the county is made up of single-family detached homes---an estimated 10,984 units. Another third of rental housing is multifamily housing located in buildings with 5 to 19 units. These are typically low-rise and garden style apartments. The remaining third is a mix of smaller multifamily like duplexes, townhomes, and manufactured homes.

The distribution of rental housing type has largely remained the same since 2010. The major increases in rental housing have been among single-family detached home rentals (2,502 added since 2010) and multifamily buildings containing 5 to 9 units (1,545 added). Although single-family rentals provide growing families with much needed space, multifamily development can more effectively address rapidly growing demand at scale.

```{r}
#| label: fig-structure
#| fig-cap: "Rental housing units by structure type"

years <- 2010:2020

b25032_vars <- load_variables(2010, "acs5") |> 
  filter(str_sub(name, end = 7) %in% "B25032_")

b25032_raw <- map_dfr(years, function(yr){
  b25032_pull <- get_acs(
    geography = "county",
    county = "Chesterfield County",
    state = "VA",
    table = "B25032",
    year = yr,
    survey = "acs5",
    cache_table = TRUE
  ) %>%
    mutate(year = yr)
  })

b25032_vars_cleaned <- b25032_vars |> 
  separate(label, into = c("est", "total", "tenure", "structure"), sep = "!!") |> 
  select(variable = name, tenure, structure) |> 
  drop_na() |> 
  filter(tenure == "Renter-occupied housing units")

b25032_data <- b25032_raw |> 
  right_join(b25032_vars_cleaned, by = "variable") |> 
  select(NAME, fips = GEOID, year, structure, estimate, moe) |> 
  mutate(structure = fct_relevel(structure,
                                 levels = c(
                                   "1, detached",
                                   "1, attached",
                                   "2",
                                   "3 or 4",
                                   "5 to 9",
                                   "10 to 19",
                                   "20 to 49",
                                   "50 or more",
                                   "Mobile home",
                                   "Boat, RV, van, etc."
                                 )))

structure_plot <- gg_bar_col(
  b25032_data,
  x_var = year,
  y_var = estimate,
  col_var = structure,
  stack = TRUE,
  pal = pals::cubicl(10),
  title = "Rental housing units by structure type",
  col_title = "Units in structure",
  col_labels = function(x) x,
  x_title = element_blank()
)

structure_plot

# plotly::ggplotly(structure_plot) %>% 
#   plotly_camera()


```

### Number of bedrooms

Most rental housing in the county are two-bedroom and three-bedroom homes. This accounts for nearly half of all rental housing in the county (68 percent). 

The number of four-bedroom rental homes has been on a steady increase since 2010---from 1,995 to 3,453 in 2020, a 73 percent increase. Although the increasing number of families in search of larger homes has led to a corresponding demand in larger rental homes, the impact can be felt among single adults, especially low-income adults, needing affordable rental opportunities. 

```{r}
#| label: fig-bedrooms
#| fig-cap: "Rental housing by bedrooms"

years <- 2010:2020

b25042_vars <- load_variables(2010, "acs5") |> 
  filter(str_sub(name, end = 6) %in% "B25042")

b25042_raw <- map_dfr(years, function(yr){
  b25042_pull <- get_acs(
    geography = "county",
    county = "Chesterfield County",
    state = "VA",
    table = "B25042",
    year = yr,
    survey = "acs5",
    cache_table = TRUE
  ) %>%
    mutate(year = yr)
  })

b25042_vars_cleaned <- b25042_vars |> 
  separate(label, into = c("est", "total", "tenure", "bedrooms"), sep = "!!") |> 
  select(variable = name, tenure, bedrooms) |> 
  drop_na() |> 
  filter(tenure == "Renter occupied")

b25042_data <- b25042_raw |> 
  right_join(b25042_vars_cleaned, by = "variable") |> 
  select(NAME, fips = GEOID, year, bedrooms, estimate, moe)

bedroom_plot <- gg_bar_col_facet(
  b25042_data,
  stack = TRUE,
  x_var = year,
  y_var = estimate,
  col_var = bedrooms,
  facet_var = bedrooms,
  y_breaks_n = 5,
  title = "Rental housing by bedrooms",
  x_title = element_blank()) +
    theme(
      legend.position = "none"
    )

bedroom_plot

# plotly::ggplotly(bedroom_plot) %>% 
#   plotly_camera()


```

### Rental housing age

The majority of existing rental housing in the county was built between 1960 and 1999. This represents a significant portion of rental housing that is at least 20 years of age. Aging rental homes often mean increasing maintenance costs or the need for significant renovations that get passed down to tenants.

```{r}
#| label: fig-yearbuilt
#| fig-cap: "Rental housing by year built"

b25036_vars <- load_variables(2020, "acs5") |> 
  filter(str_sub(name, end = 6) %in% "B25036")

b25036_raw <- get_acs(
    geography = "county",
    county = "Chesterfield County",
    state = "VA",
    table = "B25036",
    year = 2020,
    survey = "acs5",
    cache_table = TRUE
  )

b25036_vars_cleaned <- b25036_vars |> 
  separate(label, into = c("est", "total", "tenure", "yrbuilt"), sep = "!!") |> 
  select(variable = name, tenure, yrbuilt)

b25036_data <- b25036_raw |> 
  right_join(b25036_vars_cleaned, by = "variable") |> 
  select(NAME, fips = GEOID, yrbuilt, tenure, estimate, moe) |> 
  drop_na() |> 
  filter(tenure == "Renter occupied:") |> 
  mutate(yrbuilt = case_when(
    yrbuilt == "Built 2014 or later" ~ "2010 or later",
    yrbuilt == "Built 2010 to 2013" ~ "2010 or later",
    yrbuilt == "Built 2000 to 2009" ~ "2000 to 2009",
    yrbuilt == "Built 1990 to 1999" ~ "1980 to 1999",
    yrbuilt == "Built 1980 to 1989" ~ "1980 to 1999",
    yrbuilt == "Built 1970 to 1979" ~ "1960 to 1979",
    yrbuilt == "Built 1960 to 1969" ~ "1960 to 1979",
    yrbuilt == "Built 1950 to 1959" ~ "1940 to 1959",
    yrbuilt == "Built 1940 to 1949" ~ "1940 to 1959",
    yrbuilt == "Built 1939 or earlier" ~ "1939 or earlier"
  ))

built_plot <- gg_bar(
  b25036_data,
  x_var = yrbuilt,
  y_var = estimate,
  y_breaks_n = 5,
  title = "Rental housing by year built",
  x_title = element_blank()) +
    theme(
      legend.position = "none"
    )

built_plot

# plotly::ggplotly(built_plot) %>% 
#   plotly_camera()

```

## Special types of rental housing

Two significant components of Chesterfield County's rental housing supply, single-family rentals and manufactured home communities, deserve special attention. This section explores available data and trends for these unique types of housing.

### Single-family rentals

Single-family home rentals (SFRs) are a growing trend in the rental market as a result of lifestyle changes and affordability challenges. The COVID-19 pandemic has also had a dramatic impact on the demand for SFRs, as many individuals and families looked to avoid higher density apartments and have the space to work and learn remotely.^[https://www.nmhc.org/research-insight/research-notes/2021/an-overview-of-single-family-rentals/]

SFR in the county have been on the rise---over 28 percent growth in the past decade, according to ACS estimates. This is the most substantial increase among all rental housing types.

To obtain a better understanding of the scale and distribution of SFR across the county, parcel records can be used to cross-reference the property's address with the owner's address. Generally, these addresses will match for all owner-occupied homes. Those that do not match are most likely not occupied by the owner, which usually implies the house is being leased to other occupants.

Based on an analysis of these address in the latest parcel records from the county, there are an estimated 13,205 single-family home rentals in the county. These comprise nearly 12 percent of all single-family homes.

```{r sfr-setup}

# Pull in Chesterfield County magisterial districts

# districts_geojson <- "https://services3.arcgis.com/TsynfzBSE6sXfoLq/ArcGIS/rest/services/Administrative/FeatureServer/9/query?outFields=*&where=1%3D1&f=geojson"
# 
# districts <- read_sf(districts_geojson)

# View districts as simple map
# plot(districts$geometry)

# Pull in Chesterfield County parcel shapefile and convert polygons to points

# parcels <- st_read("data/ParcelsEnriched.shp") |>
#    st_point_on_surface()

# Remove unneeded columns and spatially join to magisterial districts

# parcels_sub <- parcels %>%
#   select(
#     id = GPIN,
#     use_code = UseCode,
#     owner_address = OwnerAddre,
#     owner_city = OwnerCity,
#     owner_state = OwnerState,
#     owner_zip = OwnerZip,
#     address = Address,
#     zip = Zip,
#     year_built = Year_Built,
#     improvement = Improvemen
#   ) |>
#   st_join(left = FALSE, districts["MagDistName"])

# Check closeness between parcel address and owner address

# parcels_sub$match_score <- stringsim(parcels_sub$owner_address, parcels_sub$address)

# Plot match scores

# ggplot(parcels_sub, aes(x = match_score)) + 
#  geom_histogram(bins = 30) + 
#  theme_minimal()

# Ensure consistency when sampling fuzzy address matches to manually inspect

# set.seed(1983)
# 
# fuzzy_matches <- parcels_sub %>%
#   filter(match_score >= 0.75 & match_score <= 0.9) %>%
#   st_drop_geometry() %>%
#   slice_sample(n = 100)

# Set owner-occupied SFH as those with match scores over 0.6 and whose address numbers match
# Set renter-occupied SFH as all other single-family parcels
# Assign all other parcels as "Multifamily" or "Other"

# parcels_ownership <- parcels_sub %>%
#   mutate(address_num = str_extract(address, "[0-9]+"),
#          owner_num = str_extract(owner_address, "[0-9]+")) %>%
#   mutate(ownership = case_when(
#     use_code == "SD" & match_score > 0.6 & address_num == owner_num ~ "Owner-occupied SFH",
#     use_code == "SD" ~ "Renter-occupied SFH",
#     use_code %in% c("CD", "CN", "DU", "QP", "RE", "TH", "TN", "TP", "100") ~ "Multifamily",
#     TRUE ~ "Other"
#   ))

# table(parcels_ownership$ownership)

# Filter parcels for only SFH

# ch_sfh <- filter(parcels_ownership, ownership %in% c("Owner-occupied SFH",
#                                                      "Renter-occupied SFH"))

# st_write(ch_sfh, "data/shp/ch_sfh.shp")

# ch_sfh |> 
#   st_drop_geometry() |> 
#   saveRDS("data/ch_sfh.rds")

# ch_sfr <- filter(ch_sfh, ownership == "Renter-occupied SFH")

# st_write(ch_sfr, "data/shp/ch_sfr.shp")


```

```{r}
#| label: tbl-sfr-summary
#| tbl-cap: "Single-family rentals in Chesterfield County"

# Total count and percent of SFR in the county

ch_sfh <- readRDS("data/ch_sfh.rds")

ch_sfh_sum <- ch_sfh |> 
  group_by(ownership) |> 
  summarise(cnt = n()) |> 
  mutate(pct = percent(cnt / sum(cnt), 1),
         cnt = comma(cnt, 0))

ch_sfh_sum |> 
  kable(
        align = "lrr",
        col.names = c("Ownership", "Number", "Percent")) |> 
  kableExtra::kable_styling(
    bootstrap_options = c("condensed", "hover", "striped")
  )

```

The map below shows the spatial distribution of SFRs across the county. They are found in nearly every community, but are most prevalent in Ettrick neighborhoods surrounding Petersburg and Colonial Heights, between Route 288 and Chippenham Parkway, the Route 1 corridor, and the Hull Street corridor.

```{r sfr-basemap, include=FALSE}

library(tigris)
library(ggspatial)
 
options(tigris_use_cache = TRUE)

ch_boundary <- counties("VA") |> 
  filter(NAME == "Chesterfield")
 
# ch_roads <- roads("VA", "Chesterfield", year = 2021)
# ch_water <- area_water("VA", "Chesterfield", year = 2021)

# table(ch_sfh$improvement == 0)

ch_sfr <- st_read("data/shp/ch_sfr.shp")

sf_use_s2(TRUE)

cf_basemap <- layer_static_mapbox(
  location = ch_boundary,
  style_id = "light-v10",
  username = "mapbox"
)

```

```{r}
#| label: fig-sfr-map
#| fig-cap: "Single-family rental homes in Chesterfield County"

sfr_map <- ggplot() +
  cf_basemap +
  geom_sf(data = ch_sfr, fill = "brown1", size = 1, shape = 21, alpha = .1) +
  labs(title = "Single-family rental homes in Chesterfield County",
       caption = "Source: County parcel records") +
  theme_void(base_size = 10)

sfr_map

```

When aggregated to the county's five magisterial districts, single-family rentals are most common in the Bermuda and Dale districts. SFRs comprise more than 13 percent of the total single-family stock in these districts, but this is still not significantly above the county average. The Midlothian district has the lowest share of SFRs at 9 percent.

```{r}
#| label: fig-sfr-magisterial
#| fig-cap: "Single-family rentals as percent of all single-family homes by magisterial district"

# Determine percent of SHF that are renter-occupied by magisterial district

ch_sfh_mag <- ch_sfh |> 
  group_by(MagDistName, ownership) |> 
  summarise(cnt = n()) |> 
  mutate(pct = cnt / sum(cnt)) |> 
  mutate(MagDistName = str_to_sentence(MagDistName)) |>
  mutate(MagDistName = str_replace_all(MagDistName, "Cloverhill", "Clover Hill")) |> 
  filter(ownership == "Renter-occupied SFH")

sfr_mag <- gg_hbar(ch_sfh_mag,
                   y_var = MagDistName,
                   x_var = pct,
                   y_reorder = TRUE,
                   title = "Single-family rentals as percent of all single-family homes by magisterial district",
                   x_title = "",
                   y_title = "",
                   x_labels = percent_format(),
                   y_labels = function(x) x)

sfr_mag

```

The majority of SFR owners (61 percent) also live in the county, while 29 percent live somewhere else in Virginia. The remaining 10 percent of single-family rental owners live outside of the state.

```{r sfr-owners, include=FALSE}

# sf_use_s2(FALSE)
# 
# zips <- zctas(year = 2010, state = "VA") |> 
#   st_point_on_surface()
# 
# chesterfield_zips <- zips[ch_boundary, ] |> 
#   st_drop_geometry()
# 
# write_rds(chesterfield_zips, "data/chesterfield_zips.rds")

chesterfield_zips <- read_rds("data/chesterfield_zips.rds")

zip_list <- chesterfield_zips$ZCTA5CE10

rental_class <- ch_sfh %>%
  filter(ownership == "Renter-occupied SFH") %>%
  mutate(owner_location = case_when(
    owner_zip %in% zip_list ~ "Owner in Chesterfield",
    owner_state == "VA" ~  "Owner in Virginia",
    TRUE ~ "Owner out of state"
  ))

rental_pct <- rental_class %>%
  group_by(owner_location) %>%
  summarize(n = n()) %>%
  ungroup() %>%
  mutate(pct = (n / sum(n)))

```

```{r}
#| label: fig-sfr-owners-map
#| fig-cap: "Location of single-family rental owners"

sfr_bar <- gg_bar_col(rental_pct,
                      x_var = owner_location,
                      y_var = pct,
                      col_var = owner_location,
                      col_legend_none	= TRUE,
                      y_title = "",
                      x_title = "",
                      x_labels = function(x) x,
                      y_labels = percent_format(),
                      title = "Location of single-family rental owners")

sfr_bar

```

### Manufactured homes

While single-family homes saw a major increase in rentals, manufactured homes experienced a significant decrease. From 791 in 2010 to 527 in 2020, the reported number of manufactured homes in the county has been cut down by a third.

This estimate from the U.S. Census Bureau's American Community Survey severely undercounts manufactured homes located in manufactured home communities (MHCs). Manufactured home communities are a valuable source of affordable housing across Virginia. Chesterfield County is home to a significant amount of MHCs---about 1,500 in total.

::: {.callout-note}

Additional details and maps of the county's manufactured home communities can be found in [Chapter 3](#part-1a-3).

:::

However, a significant issue faces manufactured home community residents. MHCs often give residents two options, 1) a resident may rent both their home and the lot on which that home is placed, or 2) a resident may own their home and rent the lot on which that home is placed.

In both cases, a resident is renting, but residents in the second case are in an especially precarious situation should a MHC owner choose to sell or redevelop their property. MHCs are increasingly at-risk of redevelopment.

In 2021, Suburban Village, a MHC along the Midlothian Turnpike, was [purchased](https://richmond.com/news/local/maryland-based-investment-firm-buys-chesterfield-mobile-home-park-for-22-5-million/article_41bd2656-0466-5643-9285-2715e88fee40.html) by a Maryland-based investment firm for $23 million. As the county's third largest MHC with nearly 220 households, residents and advocates fear for substantial lot rent increases or displacement.

In contrast to growing investment firm interest in MHCs, local affordable housing nonprofit, project:HOMES, [purchased](https://www.projecthomes.org/bermuda-estates) Bermuda Estates, a 52 unit MHC in 2021. Since the purchase, project:HOMES has made significant improvements to community infrastructure and has been working to replace distressed homes with new high-quality and energy efficient manufactured homes (with support from Virginia Tech's Virginia Center for Housing Research and Virginia Housing). project:HOMES's ultimate goal is to empower residents to take ownership of the community.

## Rental housing costs

The following data are from CoStar's multifamily database, which includes 118 existing rental properties in the county. Nine manufactured home communities are not found in this database; therefore, accurate and recent rent amounts for those properties are not available. Similarly, the single-family rental market is not tracked by CoStar, so these data do not include standalone homes for rent.

### Average asking rent

From the early 2000s, rental housing costs had generally been declining in the county. But changing trends in affordability and lifestyle have led to an increasing demand for rentals that has led to a complementary increase in rental housing costs. Since 2012, the average asking rent in the county has been on a steady increase. In Q2 2021, average asking rent in the county reached a two decade high of $1,515, when adjusting for inflation.

Average market asking rent per unit is expected to climb in the county---signalling an increasing demand, as well as the construction of higher end rental apartments.

```{r}
#| label: fig-costar-rents
#| fig-cap: "Average asking rent in Chesterfield County. Source: CoStar Group, Inc."}

# This is pulling in average rents for all existing multifamily properties in the county. For some reason, this differs from the previous average rent calculations. I'd resolve this by updating the previous visualization in Part-1B-4.

# CoStar filter: Multi-family, Chesterfield County, VA; Construction Status: Existing

# Pulled quarterly Market Asking Rent Per Unit; Inflation Adjusted, No Forecast, History: All

rent <- read_csv("data/costar_rent_existing_sep_6.csv") |>
  pivot_longer(cols = 2:3, names_to = "type", values_to = "value") |> 
  mutate(period = as.Date(as.yearqtr(period, format = "%Y Q%q"), frac = 1)) |> 
  mutate(type = str_replace_all(type, c("rent" = "Asking rent",
                                        "rent_adj" = "rent (inflation-adjusted)")))

costar_rent <-gg_line_col(
  rent,
  x_var = period,
  y_var = value,
  col_var = type,
  alpha_point = 0,
  size_line = 0.75,
  title = "Average asking rent in Chesterfield County",
  y_labels = label_dollar(),
  y_title = "Average asking rent",
  x_title = element_blank(),
  y_zero = TRUE,
  x_labels = scales::date_format(format = "%Y")
  ) +
  ggplot2::scale_x_date(limits = as.Date(c(NA, Sys.Date()))) +
  ggplot2::theme(
    legend.title = element_blank(),
    legend.position = "top",
    legend.direction = "horizontal",
    legend.justification = "center"
  )

costar_rent

```

### Rent by bedrooms

The number of bedrooms often impacts rental housing costs---more bedrooms generally equals higher rent. As of 2022 Q3, there is an approximate difference of $420 between the average rent for a 1 bedroom (\$1,336) versus a three bedroom apartment (\$1,756). Average rent for a 2 bedroom apartment was right in the middle at \$1,499.

```{r}
#| label: fig-bedroom-rents
#| fig-cap: "Average asking rent in Chesterfield County by bedroom."
#| fig-subcap: "Source: CoStar Group, Inc."

# This uses the same filter as above from CoStar.

bedroom_rent <- read_csv("data/costar_br_existing_sep_6.csv") |>
  pivot_longer(cols = 2:9, names_to = "type", values_to = "value") |>
  mutate(period = as.Date(as.yearqtr(period, format = "%Y Q%q"), frac = 1)) |> 
  mutate(rent = sub(".*_", "", type)) |> 
  mutate(rent = str_replace_all(rent, c("rent" = "Asking rent",
                                        "adj" = "rent (inflation-adjusted)"))) |> 
  filter(rent != "Asking rent") |> 
  mutate(type = substr(type,1,1)) |> 
  mutate(type = str_c(type, " bedroom")) |> 
  filter(type != "0 bedroom")

costar_br_rent <-gg_line_col(
  bedroom_rent,
  x_var = period,
  y_var = value,
  col_var = type,
  alpha_point = 0,
  size_line = 0.75,
  title = "Average asking rent in Chesterfield County by bedroom",
  y_labels = label_dollar(),
  y_title = "Average asking rent (inflation-adjusted)",
  x_title = element_blank(),
  y_zero = TRUE,
  x_labels = scales::date_format(format = "%Y")
  ) +
  ggplot2::scale_x_date(limits = as.Date(c(NA, Sys.Date()))) +
  ggplot2::theme(
    legend.title = element_blank(),
    legend.position = "right",
    legend.direction = "vertical",
    legend.justification = "left"
  )

costar_br_rent

```

## Rental housing construction

Like most of the country, the Great Recession negatively impacted home building in the county. That impact was most felt among single-family home construction, however. Pre-recession, the number of single-family residential construction permits averaged 2,050 per year in Chesterfield County. This pace would not be seen again until 2020.

### Multifamily building permits

Multifamily (five or more units) home starts were not as impacted by the recession. Multifamily building permits saw a substantial increase in 2019 when it reached a two decade high of 1,466. 

Duplexes and smaller multifamily homes like tri-plexes and quads have not seen substantial construction in the county. These types of properties often make up what is called "[Missing Middle](https://missingmiddlehousing.com/)" housing, a term used to describe smaller multifamily and clustered single-family housing that often offers affordability by design (i.e., smaller homes typically mean lower price).

```{r}
#| label: fig-permits
#| fig-cap: "Annual building permits in Chesterfield County by structure type"

years <- 2000:2020

header_rows <- read_csv("https://www2.census.gov/econ/bps/County/co2020a.txt", 
                        col_names = FALSE,
                        n_max = 2)

column_names <- header_rows %>%
  select(X1:X18) %>%
  t() %>%
  as_tibble() %>%
  mutate(group = rep(1:6, each = 3)) %>%
  group_by(group) %>%
  fill(V1, .direction = "updown") %>%
  mutate(names = paste0(V1, ": ", V2)) %>%
  pull(names)

cbps_raw <- map_df(years, ~{
  raw <- read_csv(glue("https://www2.census.gov/econ/bps/County/co{.x}a.txt"), skip = 2, 
                    col_names = FALSE) %>%
    select(X1:X18) %>%
    set_names(column_names)
  
  raw
  
})

cbps_data <- cbps_raw %>% 
  mutate(year = `Survey: Date`,
         GEOID = paste0(`FIPS: State`, `FIPS: County`)) %>%
  select(`1-unit: Bldgs`:GEOID) %>%
  filter(GEOID == "51041") %>%
  pivot_longer(`1-unit: Bldgs`:`5+ units: Value`,
               names_to = "type",
               values_to = "value") %>%
  separate(type, into = c("Type", "col"), sep = ": ") %>%
  pivot_wider(names_from = col,
              values_from = value) %>%
  rename_with(tolower, Type:Value) %>% 
  select(GEOID, year, type:units)

permits <- gg_bar_col_facet(
  cbps_data,
  x_var = year,
  y_var = units,
  col_var = type,
  facet_var = type,
  x_title = "",
  y_title = "",
  col_title = "",
  title = "Annual building permits in Chesterfield County by structure type")

permits

```

### Proposed and in-progress developments

There are currently 19 proposed or under construction multifamily developments in the county---an estimated total of 5,039 rental homes^[Three properties do not have number of units listed and are slated for a 2027 completion date]. Six of these properties are expected to be completed in 2022, while the remaining homes will be completed over the next two years.

```{r}
#| label: fig-costar-construction
#| fig-cap: "Upcoming rental units in Chesterfield County by delivery year"

under_con <- read_csv("data/costar_under_con_may_13.csv") |> 
  clean_names()

# sum(under_con$number_of_units, na.rm = TRUE)

under_con_yr <- under_con |> 
  filter(!is.na(number_of_units)) |> 
  group_by(year_built) |> 
  summarise(n = sum(number_of_units)) |> 
  mutate(year_built = as.character(year_built))

proposed <- gg_bar(
  under_con_yr,
  x_var = year_built,
  y_var = n,
  x_title = "",
  y_title = "",
  title = "Upcoming rental units in Chesterfield County by delivery year")

proposed

```

