# COUNTYWIDE MARKET ANALYSIS {.unnumbered}

# Market comparison {#part-2a-3}

This section will make comparisons between recent multifamily developments and existing rental supply in the county. For the purposes of this section, "recent" developers are those built in the past five years, are under construction, or are proposed. This analysis does not include single-family rentals, nor manufactured home communities.

```{r setup}

library(tidyverse)
library(zoo)
library(janitor)
library(simplevis)
library(ggplot2)
library(plotly)
library(leaflet)
library(htmlwidgets)

```

## Price

For much of the past five years, the difference in rental prices between recent and existing rental properties has remained at nearly \$400. But when the pandemic hit in the second quarter of 2020, the gap began to narrow.

Both recent and existing properties experienced a jump in average rent in the second quarter of 2020. While recent properties experienced a 1% increase in average rents, existing property average rents increased by 4% --- closing the gap to \$352. This difference in percent of change over time has continued to Q2 2022.

From Q2 2020 to Q2 2022, average rent for recent properties has changed 3% --- from \$1,696 to \$1,743. Existing property average rent growth was three times (9%) that of recent properties --- going from \$1,344 to \$1,460. From Q2 2017 to Q2 2022, the difference in average rent between recent and existing properties has gone from \$410 to \$285. 

While average rents for recent properties has remained steady, existing property rents have risen dramatically. The COVID-19 pandemic has significantly impacted demand among the existing stock of rental housing in the county --- leading to complementary increases in rental prices. 


```{r price}

# CoStar Recent Filter: Multi-family; Chesterfield County, VA; Construction Status: Existing, Under Construction, Proposed; Style: Townhome, Garden, Low-Rise, Mid-Rise, Hi-Rise; Min Year Built: 2017

# Load in costar_rent_recent.csv; Average rents for all units and by bedrooms

recent_price <- read_csv("data/costar_rent_recent.csv") |> 
  mutate(market = "recent")

# CoStar Past Filter: Multi-family; Chesterfield County, VA; Construction Status: Existing, Under Construction, Proposed; Style: Townhome, Garden, Low-Rise, Mid-Rise, Hi-Rise; Max Year Built: 2016

# Manually remove: 7626 Sandler Dr; 3201 Lady Marian Ln; 21203 Orange Hill Ave --- before conducting analytics

# Load in costar_rent_past.csv; Average rents for all units and by bedrooms

past_price <- read_csv("data/costar_rent_past.csv") |> 
  mutate(market = "existing")

# Both files cover 2017 Q1 -> 2022 Q2

# Filter "studio" out for both

price_comp <- rbind(recent_price, past_price)

avg_rent_all <- price_comp |> 
  filter(bedrooms == "All") |> 
  mutate(rent_adj = parse_number(rent_adj),
         rent = parse_number(rent),
         period = as.yearqtr(period))

avg_rent_br <- price_comp |> 
  filter(bedrooms != "Studio") |> 
  filter(bedrooms != "All") |> 
  mutate(rent_adj = parse_number(rent_adj),
         rent = parse_number(rent),
         period = as.yearqtr(period))

total_avg_rent <- gg_line_col(
  avg_rent_all,
  x_var = period,
  y_var = rent_adj,
  col_var = market)

br_avg_rent <- gg_line_col_facet(
  avg_rent_br,
  x_var = period,
  y_var = rent_adj,
  col_var = market,
  facet_var = bedrooms)

plotly::ggplotly(total_avg_rent) |> 
  plotly_camera()

```
*Source: CoStar Group, Inc.*

Price difference between recent and existing properties by number of bedrooms follows similar trends. But average rents for recent three-bedroom rentals experienced a significant increase in the middle of 2021 --- going from \$2,152 in Q2 2021 to \$2,337 in Q4 2021. In 2022, these rents began to decline somewhat, but still remained above \$2,000.

In all cases, the percent change in average rents for existing rental properties was three times that of recent properties no matter the bedroom count. One- and two-bedroom existing rentals had a near 20 percent increase in average rents from 2017 to 2022, while three-bedroom average rents increased by 16 percent. Average rents for recent rental properties only had an increase of 5 to 7 percent. The difference in rents between recent and existing properties as of Q2 2022 ranges from \$204 for one-bedrooms to \$552 for three-bedrooms.
```{r br-rent}

plotly::ggplotly(br_avg_rent) |> 
  plotly_camera()

```
*Source: CoStar Group, Inc.*

## Unit mix

Consumer preferences can change over time and those preference changes can impact market demand and prices. Multifamily building styles built in the late 20th century were largely in the garden-style apartments. The garden-style emphasizes lower densities and clustered development. Residents typically have ground floor access with no neighbors above. 

Existing rental properties are majority garden-style (85 percent). But the composition of recent multifamily development is much more diverse. Although though garden-style apartments are still being produced, the majority (82 percent) of newer multifamily properties are low- or mid-rise buildings. 

The shift towards these types of developments provides developers with greater economies of scale. More density makes it more financially possible to finance and develop properties. Without the density that comes through low- and mid-rise buildings, developers have to rely heavily on higher rents or substantial subsidies to make development possible.

``` {r unit-mix}

# Need list of existing, proposed, or under construction properties with... 

#CoStar Filters: Multi-family; Chesterfield County, VA; Construction Status: Existing, Under Construction, Proposed; Style: Townhome, Garden, Low-Rise, Mid-Rise, Hi-Rise

#Updated 7420 Ashlake Pky with Name; Updated Maisonettes Apartments with 1966 Year Built; Updated Kenswick Apartments with 1984 Year Built; Updated 3101 Danzler Cir with Name

# - Number of units
# - Style (mid-rise, garden, etc)
# - Number of 1, 2, 3, and 4 bedroom units
# - Avg sqft for 1, 2, and 3 bedroom units

unit_mix <- read_csv("data/costar_unit_mix.csv") |> 
  clean_names()

#Remove single-family homes incorrectly categorized as Low-Rise; select variables needed
unit_mix_clean <- unit_mix  |> 
  filter(property_address != "3201 Lady Marian Ln") |> 
  filter(property_address != "7626 Sandler Dr") |> 
  filter(property_address != "21203 Orange Hill Ave") |>
  filter(property_address != "12020 Winfree St") |> 
  select("address" = property_address, "name" = property_name, "status" = building_status, year_built, style, "units" = number_of_units, avg_unit_sf, rent_type, "units0" = number_of_studios_units, "units1" = number_of_1_bedrooms_units, "units2" = number_of_2_bedrooms_units, "units3" = number_of_3_bedrooms_units, "units4" = number_of_4_bedrooms_units, "sf0" = studio_avg_sf, "sf1" = one_bedroom_avg_sf, "sf2" = two_bedroom_avg_sf, "sf3" = three_bedroom_avg_sf, "sf4" = four_bedroom_avg_sf, latitude, longitude)

unit_mix_clean <- unit_mix_clean |> 
  mutate(market = case_when(
    year_built <= 2016 ~ "Existing",
    year_built >= 2017 ~ "Recent",
    name == "Ashlake Crossing" ~ "Recent"))

# BY STYLE

style_mix <- unit_mix_clean |> 
  group_by(market) |> 
  count(style)
  
style_mix <- transform(style_mix,
                       percent = ave(n,
                                     market,
                                     FUN = prop.table))

# Percent of total properties by style.

style_mix_bar <- gg_bar_col(style_mix,
                            x_var = market,
                            y_var = percent,
                            col_var = style,
                            stack = TRUE)

plotly::ggplotly(style_mix_bar) |> 
  plotly_camera()

```
*Source: CoStar Group, Inc.*

Existing rental development mainly consists of two-bedroom rental homes (55 percent) and one-bedrooms (27 percent). More recent developments have shifted towards to one-bedroom rentals (53 percent). The percent of larger rentals - two- and three-bedrooms - being produced has dropped to 36 percent and 9 percent, respectively. 

One-bedroom rentals meets a growing demand among young professionals and aging seniors in the county, but the lack of larger rental homes leaves growing families out of the multifamily rental market. Single-family homes are increasingly meeting the demand for larger families in need of rentals. Although this often is preferred by families who want greater privacy and more in-home separation, the single-family rental market can impact opportunities for potential homebuyers.

```{r br-mix}
# BY BEDROOM
br_mix_long <- unit_mix_clean |> 
  select(address, market, units0, units1, units2, units3, latitude, longitude) |> 
  pivot_longer(
    c(units0, units1, units2, units3),
    names_to = "bedrooms",
    values_to = "units"
  )

br_mix <- br_mix_long |> 
  group_by(market, bedrooms) |> 
  na.omit() |> 
  summarise(units = sum(units))


br_mix <- transform(br_mix, 
                    percent = ave(units,
                                  market,
                                  FUN = prop.table))

# Percent of total units by bedroom by market.
br_mix_bar <- gg_bar_col(br_mix,
           x_var = market,
           y_var = percent,
           col_var = bedrooms,
           stack = TRUE)

plotly::ggplotly(br_mix_bar) |> 
  plotly_camera()
```
*Source: CoStar Group, Inc.*


```{r sf-mix}
# BY SQUARE FOOTAGE

sf_mix <- unit_mix_clean |> 
  select(address, market, sf0, sf1, sf2, sf3) |> 
  pivot_longer(
    c(sf0, sf1, sf2, sf3),
    names_to = "bedroom",
    values_to = "sf"
  ) |> 
  na.omit() |> 
  mutate(bedroom = case_when(
    bedroom == "sf0" ~ "Studio",
    bedroom == "sf1" ~ "One Bedroom",
    bedroom == "sf2" ~ "Two Bedroom",
    bedroom == "sf3" ~ "Three Bedroom"
  )) |> 
  filter(bedroom != "Studio")

# Histogram of average sf of bedrooms by properties.
sf_mix_hist <- gg_histogram_col_facet(sf_mix,
                       x_var = sf,
                       col_var = market,
                       facet_var = bedroom)

plotly::ggplotly(sf_mix_hist) |> 
  plotly_camera()

```

*Source: CoStar Group, Inc.*

- BAR CHART: PERCENT OF UNITS BY STYLE (COLORED OR FACETED BY RECENT/EXISTING)
- BAR CHART: PERCENT OF TOTAL UNITS BY BEDROOM (COLORED OR FACETED BY RECENT/EXISTING)
- BAR CHART: AVERAGE SIZE (SQFT) OF UNITS BY BEDROOM (COLORED OR FACETED BY RECENT/EXISTING) 
- 

## Amenities

Compare presence of common amenities:
- "Clubhouse"
- "Property Manager on Site"
- "Pool"
- "Fitness Center"

``` {r amenities}

# Need list of existing, proposed, or under construction properties with "Amenities" field
# Create new columns for each amenity that will show T/F
# If amenity is listed in original CoStar field, its respective column will be TRUE
# Use stringr::str_detect()
# Example: mutate(pool = str_detect(amenities, "Pool"))

# May need to up the number of amenities a bit. It's not telling a great story based on the data. Mainly because there are several NULL values for recent developments because amenities are not listed yet.

amenities <- read_csv("data/costar_unit_mix.csv") |> 
  clean_names()

amenities_clean <- amenities|> 
  filter(property_address != "3201 Lady Marian Ln") |> 
  filter(property_address != "7626 Sandler Dr") |> 
  filter(property_address != "21203 Orange Hill Ave") |>
  filter(property_address != "12020 Winfree St") |> 
  select("address" = property_address, "name" = property_name, "status" = building_status, style, "units" = number_of_units, avg_unit_sf, year_built, rent_type, amenities, latitude, longitude) |> 
  mutate(market = case_when(
    year_built <= 2016 ~ "Past",
    year_built >= 2017 ~ "Recent",
    name == "Ashlake Crossing" ~ "Recent"))


amenities_clean <- amenities_clean |> 
  mutate(pool = str_detect(amenities, "Pool"),
         clubhouse = str_detect(amenities, "Clubhouse"),
         fitness = str_detect(amenities, "Fitness Center"),
         propmanager = str_detect(amenities, "Property Manager on Site"))

amenity_count <- amenities_clean |> 
  select(address, market, pool, clubhouse, fitness, propmanager) |> 
  na.omit() |> 
  rowwise() |> 
  mutate(count = as.character(sum(c_across(pool:propmanager) == "TRUE")))

amenity_summary <- amenity_count |> 
  group_by(market) |> 
  count(count)


```

- BAR CHART: PERCENT OF UNITS BY PRESENCE OF AMENITY (COLORED OR FACETED BY RECENT/EXISTING)

## Location

Copy / re-state what is already done in 2a-2?

Describe locations/neighborhoods where newer development is most common

``` {r location}

comp_map2 <- unit_mix_clean |> 
  mutate(market = case_when(
    year_built <= 2016 ~ "Existing",
    year_built >= 2017 ~ "Recent",
    name == "Ashlake Crossing" ~ "Recent")) |> 
  select(address, name, status, year_built, market, style, units, rent_type, latitude, longitude)

pal <- colorFactor(c("red", "blue"), domain = c("Recent", "Existing"))

# comp_location_map <-leaflet(comp_map2) |>
#   addTiles() |>
#   addCircleMarkers(
#     radius = 5,
#     color = ~pal(market),
#     stroke = FALSE, fillOpacity = 0.5
#    )
# 
# saveWidget(comp_location_map, "maps/comp_location_map.html")

knitr::include_url("maps/comp_location_map.html", height = "500px")

```

- MAP OF RECENT DEVELOPMENTS (COLORED BY EXISTING, UNDER CONSTRUCTION, AND PROPOSED)
- BAR CHART: PERCENT OF UNITS BY MAGISTERIAL DISTRICT (COLORED OR FACETED BY RECENT/EXISTING)

### Proximity to public transportation

Compare location of new multifamily development versus older multifamily relative to access to public transportation

- Within or outside of 1/2 mile buffer from GRTC stop

Within the county, there are only 20 out of 115 multifamily properties located within a half mile walking distance of a public transit stop. The majority of these properties are existing properties, while only six are recent developments (built in 2017 or later). 

These transit stops are mainly located at the city-county border along Midlothian Turnpike or along Route One, where the Greater Richmond Transit Company (GRTC) began local bus service from Falling Creek to John Tyler Community College in March 2020.

One recent development (Commonwealth Apartments built in 2022) has taken advantage of an Express Service Park-N-Ride located at Commonwealth Centre Parkway. This transit service provides a direct route from Commonwealth Centre Parkway to Downtown Richmond --- whether for work or leisure.


```{r transit}

library(tidyverse)
library(sf)
library(mapview)
library(crsuggest)
library(tigris)
options(tigris_use_cache = TRUE)

# # Let's take the CoStar properties as an example; the same workflow
# # will apply for the other properties as well.  
# # First, we convert to a spatial object: 
# comp_map_sf <- comp_map2 |>
#   st_as_sf(coords = c("longitude", "latitude"),
#            crs = 4326)
# 
# # Map the properties:
# mapview(comp_map_sf)
# 
# # Identify an appropriate coordinate reference system:
# va_crs <- suggest_top_crs(comp_map_sf, units = "m",
#                           inherit_gcs = FALSE)
# 
# # We'll use the Virginia South state plane CRS (6594)
# comp_prj <- st_transform(comp_map_sf, va_crs)
# 
# # Grab VA Census tracts and identify those that overlap the properties:
# chesterfield_tracts <- tracts("VA", "Chesterfield", year = 2021) |>
#   st_transform(va_crs)
# 
# costar_tracts <- st_filter(chesterfield_tracts, comp_prj)
# 
# # Options for computing proximity include:
# # - "As the crow flies" straight-line distances, which are handled in sf;
# # - Network distances / travel-times, which are easiest to compute using
# # - my mapboxapi R package.
# 
# # Transportation data (GRTC stops):
# stops <- st_read("data/grtc/Stops_May2022.shp") %>%
#   st_transform(va_crs)
# 
# # Take a quick look:
# mapview(stops)
# 
# # We can use st_distance() to compute the distance between properties and stops,
# # then find the minimum distance to determine the nearest stop
# costar_to_stops <- st_distance(comp_prj, stops)
# 
# costar_stop_distance <- apply(costar_to_stops, 1, min) %>%
#   magrittr::divide_by(1609.34)
# 
# # Alternatively, we can compute a 0.5 mile buffer around properties
# # and either count up bus stops or identify whether a bus stop is
# # accessible
# costar_buffer <- st_buffer(comp_prj, 804.67)
# 
# mapview(costar_buffer)
# 
# # Travel-times are another way to look at this.  We'll use my mapboxapi R package
# # to calculate network buffers (also called "isodistances"), travel-time matrices,
# # and also travel-time isochrones.
# #
# # I'm working on a couple updates to the package so we'll need the
# # development version from GitHub:
# remotes::install_github("walkerke/mapboxapi")
# library(mapboxapi)
# 
# # You'll need a Mapbox account + an access token to use its services.
# # Head over to https://account.mapbox.com/ and sign up for the service, then
# # grab your default access token.  Set it in R as follows, and install it
# # if you'd like:
# mb_access_token(MAPBOX_PUBLIC_TOKEN)
# 
# # You'll now be able to compute network buffers and distance/travel-times
# # based on a road network and typical traffic (or a given travel mode).
# # costar_network_buffer <- mb_isochrone(costar_prj, profile = "walking",
# #                                       distance = 805)
# #
# # mapview(costar_network_buffer)
# # Note how different the 0.5 mile network buffers are from the straight-line buffers.
# 
# # To get the shortest walking distance, we can compute a travel-distance matrix
# # Due to internal limitations of the Mapbox matrix API, the volume of stops, and the increased complexity of building network-based matrices
# # this will be slow to get the dense matrix.  This took me a few minutes to compute.
# # costar_stop_walking <- costar_prj %>%
# #   mutate(chunk = ntile(n = 2)) %>%
# #   split(~chunk) %>%
# #   map(~{
# #     mb_matrix(origins = .x,
# #               destinations = stops,
# #               profile = "walking",
# #               output = "distance")
# #   }) %>%
# #   reduce(rbind) %>%
# #   apply(1, min) %>%
# #   magrittr::divide_by(1609.34)
# 
# # A (much!) faster way to handle this is be to iterate through the apartments, find the nearest n stops by straight-line distance, then calculate the network distance to them.  That would look like this for the nearest 20 stops:
# library(nngeo)
# 
# nearest_stops <- st_nn(comp_prj, stops, k = 20)
# 
# costar_stop_walking <- imap(nearest_stops, ~{
#   # Get the specific property (the list index)
#   property <- comp_prj[.y,]
# 
#   # Get the nearby stops (the indices in the list element)
#   nearby_stops <- stops[.x, ]
# 
#   # Calculate the matrix
#   mb_matrix(origins = property,
#             destinations = nearby_stops,
#             profile = "walking",
#             output = "distance")
# }) %>%
#   reduce(rbind) %>%
#   apply(1, min) %>%
#   magrittr::divide_by(1609.34)
# 
# # Interestingly, the walking distances are reasonably different from the straight-line distances,
# # and sometimes far exceed them.
# 
# # From here, you can readily append both the minimum walking distance to
# # a transit stop to the CoStar data, then determine which properties are
# # within 0.5 miles of a transit stop
# comp_map_sf <- comp_map_sf |>
#   mutate(
#     walk_to_transit = costar_stop_walking,
#     transit_nearby = ifelse(walk_to_transit <= 0.5, 1, 0)
#   )
# 
# comp_transit <- comp_map_sf |>
#   filter(transit_nearby >= 1)

# pal <- colorFactor(c("red", "blue"), domain = c("Recent", "Existing"))
#
# transit_map <-leaflet(comp_transit) |>
#   addTiles() |>
#   addCircleMarkers(
#     radius = 5,
#     color = ~pal(market),
#     stroke = FALSE, fillOpacity = 0.5
#    )
#
# saveWidget(transit_map, "maps/transit_map.html")

knitr::include_url("maps/transit_map.html", height = "500px")

```
### Proximity to schools

Compare location of new multifamily development versus older multifamily relative to access to public schools

- Within or outside of 30 minute drive time


```{r schools}
# Once this workflow is settled, you can reasonably reproduce it for the other datasets.  
# schools <- st_read("https://services3.arcgis.com/TsynfzBSE6sXfoLq/ArcGIS/rest/services/Administrative/FeatureServer/15/query?outFields=*&where=1%3D1&f=geojson") %>%
#   st_transform(va_crs)
# 
# nearest_schools <- st_nn(comp_prj, schools, k = 20)
# 
# costar_school_driving <- imap(nearest_schools, ~{
#   # Get the specific property (the list index)
#   property <- comp_prj[.y,]
#   
#   # Get the nearby schools (the indices in the list element)
#   nearby_schools <- schools[.x, ]
#   
#   # Calculate the matrix (drive-times here instead)
#   # Minutes are returned by default
#   # There is also a "driving-traffic" profile but unfortunately 
#   # the Matrix API doesn't allow you to specify a time of day, so it 
#   # gives you a mix of historical and live traffic instead.
#   mb_matrix(origins = property,
#             destinations = nearby_schools,
#             profile = "driving",
#             output = "duration")
# }) %>%
#   reduce(rbind) %>%
#   apply(1, min)
# 
# comp_map_sf <- comp_map_sf |> 
#   mutate(
#     minutes_to_school = costar_school_driving,
#     school_nearby = ifelse(minutes_to_school <= 30, 1, 0))

# All properties are within a distance of at least one school. Maybe further analysis required to determine what time of school?



```
### Proximity to grocery stores

Compare location of new multifamily development versus older multifamily relative to access to grocery stores

- Within or outside of 1/2 mile buffer from grocery store

```{r grocery}
# Grocery stores: 
# We'll use SNAP retailers from the USDA as a proxy here
# This is not perfect so some filtering may be necessary
# as it picks up on convenience stores as well
# snap <- read_csv("https://opendata.arcgis.com/datasets/e9cc76a48ccb45628181ece7b2deb56d_0.csv") %>%
#   filter(State == "VA") %>%
#   st_as_sf(coords = c("X", "Y"), crs = 4326) %>%
#   st_transform(va_crs)
# 
# nearest_snap <- st_nn(comp_prj, snap, k = 20)
# 
# 
# comp_grocery <- imap(nearest_snap, ~{
#   # Get the specific property (the list index)
#   property <- comp_prj[.y,]
#   
#   # Get the nearby stops (the indices in the list element)
#   nearby_snap <- snap[.x, ]
#   
#   # Calculate the matrix
#   mb_matrix(origins = property,
#             destinations = nearby_snap,
#             profile = "walking",
#             output = "distance")
# }) %>%
#   reduce(rbind) %>%
#   apply(1, min) %>%
#   magrittr::divide_by(1609.34)
# 
# comp_map_sf <- comp_map_sf |> 
#   mutate(
#     miles_to_snap = comp_grocery,
#     snap_nearby = ifelse(miles_to_snap <= 0.5, 1, 0))
# 
# comp_snap <- comp_map_sf |> 
#   filter(snap_nearby >= 1)

# pal <- colorFactor(c("red", "blue"), domain = c("Recent", "Existing"))
# 
# snap_map <- leaflet(comp_snap) |> 
#   addTiles() |> 
#   addCircleMarkers(
#     radius = 5,
#     color = ~pal(market),
#     stroke = FALSE, fillOpacity = 0.5
#    )
# 
# saveWidget(snap_map, "maps/snap_map.html")

knitr::include_url("maps/snap_map.html", height = "500px")

```

### Proximity to healthcare

Compare location of new multifamily development versus older multifamily relative to access to hospitals and urgent care facilities

- Within or outside of 15 minute drive time


```{r medical}
# Hospital / urgent care
# I'm not filtering by type but you might consider doing so
# There are some other datasets in HIFLD on care centers that may 
# be relevant as well.
# hospitals <- st_read("https://services1.arcgis.com/Hp6G80Pky0om7QvQ/arcgis/rest/services/Hospitals_1/FeatureServer/0/query?outFields=*&where=1%3D1&f=geojson") %>%
#   filter(STATE == "VA") %>%
#   st_transform(va_crs)
# 
# nearest_hospitals <- st_nn(comp_prj, hospitals, k = 20)
# 
# costar_hospital_driving <- imap(nearest_hospitals, ~{
#   # Get the specific property (the list index)
#   property <- comp_prj[.y,]
#   
#   # Get the nearby hospitals (the indices in the list element)
#   nearby_hospitals <- hospitals[.x, ]
#   
#   # Calculate the matrix (drive-times here instead)
#   # Minutes are returned by default
#   mb_matrix(origins = property,
#             destinations = nearby_hospitals,
#             profile = "driving",
#             output = "duration")
# }) %>%
#   reduce(rbind) %>%
#   apply(1, min) 
# 
# comp_map_sf <- comp_map_sf |> 
#   mutate(
#     minutes_to_hospital = costar_hospital_driving,
#     hospital_nearby = ifelse(minutes_to_hospital <= 15, 1, 0))
    
# Jobs are pending in the Google Doc (though I can show you how to 
# create jobs clusters with LODES data)

# Finally, we can put all the data back together with minimum travel times
# and critical distance / duration flags.
# costar_output <- costar %>%
#   mutate(
#     miles_to_transit = costar_stop_walking,
#     transit_nearby = ifelse(miles_to_transit <= 0.5, 1, 0),
#     minutes_to_school = costar_school_driving,
#     school_nearby = ifelse(minutes_to_school <= 30, 1, 0),
#     miles_to_snap = costar_snap_walking,
#     snap_nearby = ifelse(miles_to_snap <= 0.5, 1, 0),
#     minutes_to_hospital = costar_hospital_driving,
#     hospital_nearby = ifelse(minutes_to_hospital <= 15, 1, 0)
#   )


```

### Proximity to job centers

TBD, waiting for feedback from county staff