# Rental spatial analysis

This section presents an analysis of the county’s existing rental supply, focusing on the amount and location of rental housing units. 

# Recent multifamily developments

Spatial analysis of recent multifamily developments; summarize general physical traits and location; talk about price points and contextualize with amenities or no amenities.


```{r recent}

```

# Market affordable rental

Spatial analysis of market affordable housing; summarize general physical traits and location; previously covered age and price point; pull in some information from previous but expand

```{r noah}

```


# Affordable rental

Spatial analysis of affordable housing projects; describe projects, developer, subsidy, unit mix; status of subsidy; ask Jessica for list of state-funded projects in county


```{r affordable}

```

# LIHTC

Low-income Housing Tax Credit projects are currently one of the major ways in which affordable housing is developed across the country. This program utilizes a tax incentive to 