# Rental market summary 

This section summarizes the existing stock of rental units within the county and provides an analysis of recent price and production trends in the rental market.

```{r setup}

library(tidyverse)
library(tidycensus)
library(glue)
library(esquisse)
library(plotly)
library(scales)
library(formattable)
library(simplevis)
library(lubridate)
library(patchwork)
library(knitr)
library(zoo)

```
## Rental housing stock


### Total renters
Over the past decade, the number of renter households in the county has steadily been increasing. From 2010 to 2019, the number of renters in the county has gone from 24,312 to 30,273 --- a 24 percent increase. 

In 2020, there was a two percent decrease in renters. But this decrease of 633 renters can be attributed to the pitfalls of the 2020 Census count.

```{r tenure}

years <- 2010:2020

b25003_vars <- load_variables(2010, "acs5") |> 
  filter(str_sub(name, end = 7) %in% "B25003_")

b25003_raw <- map_dfr(years, function(yr){
  b25003_pull <- get_acs(
    geography = "county",
    county = "Chesterfield County",
    state = "VA",
    table = "B25003",
    year = yr,
    survey = "acs5",
    cache_table = TRUE
  ) %>%
    mutate(year = yr)
  })

b25003_vars_cleaned <- b25003_vars |> 
  separate(label, into = c("est", "total", "tenure"), sep = "!!") |> 
  select(variable = name, tenure) |> 
  drop_na() |> 
  mutate(tenure = case_when(
    tenure == "Owner occupied" ~ "Homeowner",
    tenure == "Renter occupied" ~ "Renter"
  )) |> 
  filter(tenure == "Renter")

b25003_data <- b25003_raw |> 
  right_join(b25003_vars_cleaned, by = "variable") |> 
  select(NAME, fips = GEOID, year, tenure, estimate, moe)
  

renters_plot <- gg_line(
  b25003_data,
  x_var = year,
  y_var = estimate,
  y_breaks_n = 5,
  title = "Renter households",
  x_title = element_blank())

plotly::ggplotly(renters_plot) %>% 
  plotly_camera()

```

### Housing type

Single-family home rentals in the county have been on the rise. In 2010, 8,482 single-family detached homes were being used as rentals. By 2020, that number had increased to 10,894 --- just over a third of all rental housing in the county. This equates to a 28 percent increase.

While single-family homes saw a major increase in rentals, manufactured homes experienced a significant decrease. From 791 in 2010 to 527 in 2020, the reported number of manufactured homes in the county has been cut down by a third.

```{r structure}

years <- 2010:2020

b25032_vars <- load_variables(2010, "acs5") |> 
  filter(str_sub(name, end = 7) %in% "B25032_")

b25032_raw <- map_dfr(years, function(yr){
  b25032_pull <- get_acs(
    geography = "county",
    county = "Chesterfield County",
    state = "VA",
    table = "B25032",
    year = yr,
    survey = "acs5",
    cache_table = TRUE
  ) %>%
    mutate(year = yr)
  })

b25032_vars_cleaned <- b25032_vars |> 
  separate(label, into = c("est", "total", "tenure", "structure"), sep = "!!") |> 
  select(variable = name, tenure, structure) |> 
  drop_na() |> 
  filter(tenure == "Renter-occupied housing units")

b25032_data <- b25032_raw |> 
  right_join(b25032_vars_cleaned, by = "variable") |> 
  select(NAME, fips = GEOID, year, structure, estimate, moe)

structure_plot <- gg_bar_col(
  b25032_data,
  stack = TRUE,
  x_var = year,
  y_var = estimate,
  col_var = structure,
  y_breaks_n = 5,
  title = "Rental housing by structure type",
  x_title = element_blank()) +
    theme(
      legend.position = "none"
    )

plotly::ggplotly(structure_plot) %>% 
  plotly_camera()


```

SFRs
```{r sfr}

```


Most rental housing in the county are two-bedroom and three-bedroom homes. This accounts for nearly half of all rental housing in the county (68 percent). 

The number of four-bedroom rental homes has been on a steady increase since 2010 --- from 1,995 to 3,453 in 2020, a 73 percent increase. Although the increasing number of families in search of larger homes has led to a corresponding demand in larger rental homes, the impact can be felt among single adults, especially low-income adults, needing affordable rental opportunities. 

```{r bedrooms}

years <- 2010:2020

b25042_vars <- load_variables(2010, "acs5") |> 
  filter(str_sub(name, end = 6) %in% "B25042")

b25042_raw <- map_dfr(years, function(yr){
  b25042_pull <- get_acs(
    geography = "county",
    county = "Chesterfield County",
    state = "VA",
    table = "B25042",
    year = yr,
    survey = "acs5",
    cache_table = TRUE
  ) %>%
    mutate(year = yr)
  })

b25042_vars_cleaned <- b25042_vars |> 
  separate(label, into = c("est", "total", "tenure", "bedrooms"), sep = "!!") |> 
  select(variable = name, tenure, bedrooms) |> 
  drop_na() |> 
  filter(tenure == "Renter occupied")

b25042_data <- b25042_raw |> 
  right_join(b25042_vars_cleaned, by = "variable") |> 
  select(NAME, fips = GEOID, year, bedrooms, estimate, moe)

bedroom_plot <- gg_bar_col_facet(
  b25042_data,
  stack = TRUE,
  x_var = year,
  y_var = estimate,
  col_var = bedrooms,
  facet_var = bedrooms,
  y_breaks_n = 5,
  title = "Rental housing by bedrooms",
  x_title = element_blank()) +
    theme(
      legend.position = "none"
    )

plotly::ggplotly(bedroom_plot) %>% 
  plotly_camera()

```

CoStar does not collect data on single-family rentals, therefore it is difficult to understand the single-family rental, or SFR, market in the county. 

Across CoStar's multifamily database, however, there are 118 existing rental properties. This excludes nine manufactured home parks identified by the Manufactured Housing Community Coalition of Virginia in their 2016 Central Virginia study.

There are at least 23,377 homes across these 127 properties.


```{r costar-buildings}

existing <- read_csv("data/costar_existing.csv")

```

## Rental housing costs

-Focus on CoStar data. Maybe look at ACS 1-year rents - which includes single family rental with MOE.
 Use PUMS for single-family/townhomes - below 5 units - best estimate. 
```{r costar-rents}

# This is pulling in average rents for all existing multifamily properties in the county. For some reson, this differs from the previous average rent calculations for some reason. I'd resolve this by updating the previous visualization in Part-1B-4.


rent <- read_csv("data/costar_rent_existing_may_12.csv") |>
  pivot_longer(cols = 2:3, names_to = "type", values_to = "value") |> 
  mutate(period = as.Date(as.yearqtr(period, format = "%Y Q%q"), frac = 1)) |> 
  mutate(type = str_replace_all(type, c("rent" = "Asking rent",
                                        "rent_adj" = "rent (inflation-adjusted)")))


costar_rent <-gg_line_col(
  rent,
  x_var = period,
  y_var = value,
  col_var = type,
  alpha_point = 0,
  size_line = 0.75,
  title = "Average asking rent in Chesterfield County",
  y_labels = label_dollar(),
  y_title = "Average asking rent",
  x_title = element_blank(),
  y_zero = TRUE,
  x_labels = scales::date_format(format = "%Y")
  ) +
  ggplot2::scale_x_date(limits = as.Date(c(NA, Sys.Date()))) +
  ggplot2::theme(
    legend.title = element_blank(),
    legend.position = "top",
    legend.direction = "horizontal",
    legend.justification = "center"
  )

plotly::ggplotly(costar_rent) %>% 
  plotly_camera()



```


```{r bedroom-rents}

bedroom_rent <- read_csv("data/costar_br_existing_may_12.csv") |>
  pivot_longer(cols = 2:9, names_to = "type", values_to = "value") |>
  mutate(period = as.Date(as.yearqtr(period, format = "%Y Q%q"), frac = 1)) |> 
  mutate(rent = sub(".*_", "", type)) |> 
  mutate(rent = str_replace_all(rent, c("rent" = "Asking rent",
                                        "adj" = "rent (inflation-adjusted)")))
```

## Rental housing construction

There are currently 19 proposed or under construction multifamily 

```{r costar-construction}

```




-Outline multifamily building permit trends in county.

```{r permits}

years <- 2000:2020

header_rows <- read_csv("https://www2.census.gov/econ/bps/County/co2020a.txt", 
                        col_names = FALSE,
                        n_max = 2)

column_names <- header_rows %>%
  select(X1:X18) %>%
  t() %>%
  as_tibble() %>%
  mutate(group = rep(1:6, each = 3)) %>%
  group_by(group) %>%
  fill(V1, .direction = "updown") %>%
  mutate(names = paste0(V1, ": ", V2)) %>%
  pull(names)

cbps_raw <- map_df(years, ~{
  raw <- read_csv(glue("https://www2.census.gov/econ/bps/County/co{.x}a.txt"), skip = 2, 
                    col_names = FALSE) %>%
    select(X1:X18) %>%
    set_names(column_names)
  
  raw
  
})

cbps_data <- cbps_raw %>% 
  mutate(year = `Survey: Date`,
         GEOID = paste0(`FIPS: State`, `FIPS: County`)) %>%
  select(`1-unit: Bldgs`:GEOID) %>%
  filter(GEOID == "51041") %>%
  pivot_longer(`1-unit: Bldgs`:`5+ units: Value`,
               names_to = "type",
               values_to = "value") %>%
  separate(type, into = c("Type", "col"), sep = ": ") %>%
  pivot_wider(names_from = col,
              values_from = value) %>%
  rename_with(tolower, Type:Value) %>% 
  select(GEOID, year, type:value)


```

